---
title: "Data Analysis"
author: ""
date: ""
output:
  bookdown::word_document2:
    fig_caption: yes
    toc: no
    toc_depth: '3'
  bookdown::pdf_document2:
    latex_engine: lualatex
    toc: yes
    toc_depth: '3'
  bookdown::html_document2:
    highlight: tango
    number_sections: yes
    theme: united
    toc: no
    toc_depth: 3
---


```{r setup , include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE,
                      warning = FALSE,
                      cache = FALSE,
                      fig.align = "center",
                      fig.width = 6, 
                      fig.asp = 0.618, 
                      fig.show = "hold")

options(tinytex.verbose = TRUE)
options(dplyr.print_min = 6, dplyr.print_max = 6)
options(digits = 4)
options(scipen=999)
```

## Statistical Analysis

We present descriptive summaries of key sample statistics. We compared baseline sample characteristics between those children found to be at risk of OSA and those not at risk. We used the Fisher's exact test or chi-square tests for categorical variables. The Wilcoxon rank sum test or Kruskal–Wallis test was used for continuous variables. We considered p-values of <0.05 to be statistically significant. <br>


```{r, packages, results='hide'}

knitr::opts_chunk$set(chunk_option1 = TRUE)

rm(list = ls(all=T))
browseURL("https://stats.idre.ucla.edu/r/dae/logit-regression/")
browseURL("https://stackoverflow.com/questions/8161836/how-do-i-replace-na-values-with-zeros-in-an-r-dataframe")
browseURL("https://stackoverflow.com/questions/15111629/create-summary-table-of-categorical-variables-of-different-lengths")

list.pkg <- c("plyr", "tidyr", "rlang", "tidydplyr::select", "gWidgets","RGtk2","gWidgetsRGtk2","data.table","RCurl","zoo","RedcapData",
           "devtools", "ReporteRs","stringr", "dplyr","grid","ggplot2","zoo", "lubridate", "digest", "janitor", "gt", "ts", "magrittr", "rsample", "tidymodels", "skimr","tibble", "gridExtra", "ggforce", "colorspace", "colorblindr","cowplot","dviz.supp", "gee", "lme4", "multcomp")

ipk = function(pkg){
  new.pkg = list.pkg[!(list.pkg %in% installed.packages()[, "Package"])]
  if("practicalgg" %in% list.pkg){
    remotes::install_github("wilkelab/practicalgg")
  }
  if(length(new.pkg)) install.packages(new.pkg, dependencies = T ,repos = "http://cran.us.r-project.org")
  sapply(pkg, require, character.only = T)
}
ipk(list.pkg)
#source(file.path(home.dir,"hsuApps", "universalFunctions", "modifyData.R"))

if(tolower(Sys.info()[['sysname']])=="linux"){
  home.dir <- normalizePath(paste0(Sys.getenv()[["HOME"]])) %>% sub("\\\\", "/", .)
}else{
  home.dir <- normalizePath(paste0(Sys.getenv()[["USERPROFILE"]],"/Analysis")) %>% gsub("\\\\", "/", .)
}

proj_dir = file.path(home.dir, 'Sylvia')
data_dir = file.path(proj_dir, 'data')
cache_dir = file.path(proj_dir, "cache", "doc")
src_dir = file.path(proj_dir, "src", "r")

for(i in c(proj_dir, data_dir, cache_dir, src_dir)){
  if(!dir.exists(i)){
  if(dir.create(i, recursive = T))
    stop("Project directory\"", i, 
         "\"has been created!\nPlease fill it with relevant files")
  else
    stop("The directory", i, "could not be created")
  
  }
}

```



```{r baseline}

knitr::opts_chunk$set(chunk_option1 = TRUE)

data = readr::read_csv(file.path(data_dir, "data.csv")) %>%
  janitor::clean_names()

meta_data = readr::read_table(file.path(data_dir, "meta data.txt"))
meta_data = meta_data %>%
  dplyr::select(-grep("x", colnames(.), ignore.case = T)) %>%
  janitor::clean_names() %>%
  mutate(., name = tolower(name))

# psq before surgery
data = data %>%
  rename(stop_breathing_night_b = v71, breath_mouth_b = v72, dry_mouth_b = v73, wet_bed_b = v74)

meta_data$name = gsub('\\bv71\\b', 'stop_breathing_night_b', meta_data$name)
meta_data$name = gsub('\\bv72\\b', 'breath_mouth_b', meta_data$name)
meta_data$name = gsub('\\bv73\\b', 'dry_mouth_b', meta_data$name)
meta_data$name = gsub('\\bv74\\b', 'wet_bed_b', meta_data$name)


data = data %>%
  mutate(., hospital = ifelse(grepl('knh', hosp, ignore.case = T), 'public', 'private'),
         `birth weight` = ifelse(bth_wght < 2.5, '< 2.5 kgs', '>= 2.5 kgs'))
data$psq_b_yes = rowSums(data[,69:90] == 1, na.rm = T) # 1 = yes, 2 = no, 0 = dont know
data$psq_b = round(data$psq_b_yes/(22- rowSums(data[,69:90] == 0, na.rm = T)),2)
data$psq_b_binary = ifelse(data$psq_b > 0.33, 1,0) # 1 = psq suspected = SDBS
data$age = ifelse(!is.na(data$age_mnths), data$age_yrs + round(data$age_mnths/12,1), data$age_yrs)

# psq after surgery

data$psq_yes = rowSums(data[,37:58] == 1, na.rm = T) # 1 = yes, 2 = no, 0 = dont know
data$psq = round(data$psq_yes/(22- rowSums(data[,37:58] == 0, na.rm = T)),2)
data$psq_binary = ifelse(data$psq > 0.33, 1,0) # 1 = psq suspected = SDBS

# quantile(data$bmi, probs = c(0.05, 0.85, 0.95)) # bmi categories

data = data %>%
  mutate(., bmi_cat = ifelse(bmi < quantile(bmi, probs = c(0.05)), "Underweightt",
                             ifelse(bmi >= quantile(bmi, probs = c(0.05)) & bmi < quantile(bmi, probs = c(0.85)), "Normal",
                                    ifelse(bmi >= quantile(bmi, probs = c(0.85)) & bmi < quantile(bmi, probs = c(0.95)), "Overweight", "Obese"))))

data = data %>%
  mutate(., age_cat = ifelse(age <= 5, "2-5", "6-10"),
         asthma_diag = ifelse(child_asthma == 1 | wheezing == 1| medication_a ==1 | medication ==1 | medication_d ==1 | medication_e ==1, 1,0),
         allergic_rhinitis_diag = ifelse(allergic_rhinitis ==1 | a_symptoms == 1 | b_symptoms ==1 | c_symptoms == 1, 1, 0))

names = names(data[, 59:96])

osa_cols = c('shake_breath', 'shake_breath_b', 'apnes_b', 'apnoea_during_sleep', 'struggle_breathing', 'struggle_breath_b', 'conserns', 'concrn_breath_b', 'loud_snore_b', 'snore_loud', 'snore_sleep', 'snore_sleep_b')

data = data %>%
  mutate_at(vars(shake_breath, shake_breath_b, apnes_b, apnoea_during_sleep, struggle_breathing, struggle_breath_b, conserns, concrn_breath_b, loud_snore_b, snore_loud, snore_sleep, snore_sleep_b), ~replace(., is.na(.), 0))

# osa_cols = data[, 59:96]

for(c in osa_cols){
data[,c] = sapply(data[,c], function(x){
  x = ifelse(is.na(x), 0, x)
})
}
# severe OSA before surgery

data$A = round((data$shake_breath + data$apnoea_during_sleep)/2,2)
data$B = round((data$A + data$struggle_breathing)/2,2)
data$C = round((data$B + data$conserns)/2,2)
data$D = round((data$C + data$snore_loud)/2,2)
data$E = round((data$D + data$snore_sleep)/2,2)
data$osa_binary = ifelse(data$E >=2.72, 1, 0)

# severe osa after surgery

data$AA = round((data$shake_breath_b + data$apnes_b)/2,2)
data$BB = round((data$AA + data$struggle_breath_b)/2,2)
data$CC = round((data$BB + data$concrn_breath_b)/2,2)
data$DD = round((data$CC + data$loud_snore_b)/2,2)
data$EE = round((data$DD + data$snore_sleep_b)/2,2)
data$osa_b_binary = ifelse(data$E >=2.72, 1, 0)
# 
# tmp <- do.call(data.frame,
#            list(mean = apply(data, 2, mean),
#                 sd = apply(data, 2, sd),
#                 median = apply(data, 2, median),
#                 min = apply(data, 2, min),
#                 max = apply(data, 2, max),
#                 n = apply(data, 2, length)))
# tmp
# 
# age = data %>%
#   summarise(`mean age` = mean(age, na.rm = T),
#             `min age` = min(age, na.rm = T),
#             `max age` = max(age, na.rm = T))
# age  = age %>%
#   gather(key = "characteristic", value = "all participants")


# all charactersitics at baseline

DataXXX = data
source(file.path(src_dir, "data_clean.R"))

summ(data)

# with SDB at baseline

data_sdb = filter(data, psq_binary == 1)

summ(data_sdb)
names(out_data_sdb) = c("Characteristics", "PSQ > 0.33")

# no SDB at baseline

data_no_sdb = filter(data, psq_binary == 0)
summ(data_no_sdb)
names(out_data_no_sdb) = c("Characteristics", "PSQ < 0.33")

# with OSA at baseline

data_osa = filter(data, psq_binary ==1 & osa_binary== 1)
summ(data_osa)
names(out_data_osa) = c("Characteristics", "POSAT >= 2.72")

# no OSA at baseline

data_no_osa = filter(data, psq_binary == 1 & osa_binary== 0)
summ(data_no_osa)
names(out_data_no_osa) = c("Characteristics", "POSAT < 2.72")



sdb_baseline = left_join(out_data, out_data_sdb) %>%
  left_join(out_data_no_sdb)

#  test statistics and p-values


test_stat(data_sdb, data_no_sdb)
# data = data_sdb; data_1 =  data_no_sdb

sdb_baseline_out = sdb_baseline %>%
  left_join(., p_val_data_sdb) %>%
  left_join(., conf_data_sdb)

sdb_baseline_out = sdb_baseline_out %>%
  mutate_if(is.numeric, ~replace(., is.na(.), '')) %>%
  mutate_if(is.character,~replace(., is.na(.), ''))

names(sdb_baseline_out) = c("Characteristics", paste0('All participants\n','n=', nrow(data)), paste0("PSQ > 0.33\n", "n=", sum(data$psq_binary == 1)), paste0("PSQ < 0.33\n", "n=", sum(data$psq_binary == 0)), "p-value", "CI")

# Baseline table


options("ReporteRs-fontsize" = 12);options("ReporteRs-default-font" = "Calibri")

table.baseline <- tab_1 %>% vanilla.table(.) %>% 
  setZebraStyle(., odd = '#eeeeee', even = 'white')
table.baseline[,1] = parLeft()

table.end <- tab_2 %>% vanilla.table(.) %>% 
  setZebraStyle(., odd = '#eeeeee', even = 'white')
table.end[,1] = parLeft()

doc = docx() %>%

  # Add spaces to center the cover page
  addParagraph(., c("\n")) %>% addParagraph(., c("\n")) %>% addParagraph(., c("\n")) %>%
  

addSection(., landscape = FALSE)%>% # Change Orientation of the page
addTitle(., value = 'TABLE 1. Socio - demographic characteristics of participants at baseline') %>%
addFlexTable(., flextable = table.baseline) %>%

addParagraph(., c("\n")) %>% addParagraph(., c("\n")) %>% addParagraph(., c("\n")) %>%
  

addSection(., landscape = FALSE)%>% # Change Orientation of the page
addTitle(., value = 'TABLE 2. Socio - demographic characteristics of participants post surgery') %>%
addFlexTable(., flextable = table.end)

writeDoc(doc, file = file.path(cache_dir, paste('socio-demographic characteristics.docx')))

table.baseline <- sdb_baseline_out %>% vanilla.table(.) %>% 
  setZebraStyle(., odd = '#eeeeee', even = 'white')
table.baseline[,1] = parLeft()

doc = docx() %>%

  # Add spaces to center the cover page
  addParagraph(., c("\n")) %>% addParagraph(., c("\n")) %>% addParagraph(., c("\n")) %>%
  

addSection(., landscape = FALSE)%>% # Change Orientation of the page
addTitle(., value = 'TABLE 2. Baseline characteristics of all participants with and without SDB') %>%
addFlexTable(., flextable = table.baseline)

table.sdb <- prop_out %>% vanilla.table(.) %>% 
  setZebraStyle(., odd = '#eeeeee', even = 'white')
table.baseline[,1] = parLeft()

doc = docx() %>%

  # Add spaces to center the cover page
  addParagraph(., c("\n")) %>% addParagraph(., c("\n")) %>% addParagraph(., c("\n")) %>%
  

addSection(., landscape = FALSE)%>% # Change Orientation of the page
addTitle(., value = 'TABLE 1. Baseline characteristics of all participants with SDB') %>%
addFlexTable(., flextable = table.sdb)
writeDoc(doc, file = file.path(cache_dir, paste('SDB prevalence.docx')))


writeDoc(doc, file = file.path(cache_dir, paste('baseline characteristics.docx')))
knitr::kable(mtcars[1:10,], caption = "Baseline characteristics of all participants with and without SDB'")
```

## Descriptive Analysis

Overal, a total of `r nrow(data)` children with a mean age of `r round(mean(data$age),1)` ( `r min(data$age)`, `r max(data$age)`) were recruited. Of these, `r nrow(filter(data, sex == 1))` (`r paste0(round(nrow(filter(data, sex ==1))/ nrow(data)*100,0), "%")`) were male. Comparing baseline characteristics between children with possible diagnosis of SDB `r sum(data$psq_binary == 1)` ( `r round(sum(data$psq_binary == 1)/nrow(data)*100,2)`) and those without revealed a significant difference in  suspected allergic rhinitis (p=0.017) and those with suspected OSA (p < 0.000). There were however no significant differences in other population characteristics as shown in table \@ref(tab:baseline). <br>

```{r, endline}
knitr::opts_chunk$set(chunk_option1 = TRUE)

# with SDB at endline

data_sdb_end = filter(data, psq_b_binary == 1)
summ(data_sdb_end)
names(out_data_sdb_end) = c("Characteristics", "PSQ > 0.33")

# no SDB at endline

data_no_sdb_end = filter(data, psq_b_binary == 0)
summ(data_no_sdb_end)
names(out_data_no_sdb_end) = c("Characteristics", "PSQ < 0.33")

# with OSA at endline

data_osa_end = filter(data, psq_b_binary ==1 & osa_b_binary== 1)
summ(data_osa_end)
names(out_data_osa_end) =c("Characteristics", value = "POSAT >= 2.72")

# no OSA at endline

data_no_osa_end = filter(data, psq_b_binary == 1 & osa_b_binary== 0)
summ(data_no_osa_end)
names(out_data_no_osa_end)= c("Characteristics", value = "POSAT < 2.72")

sdb_endline = left_join(out_data, out_data_sdb_end) %>%
  left_join(out_data_no_sdb_end)

#  test statistics and p-values


test_stat_bmi_hosp(data_sdb_end, data_no_sdb_end)

# sdb_baseline_out = sdb_baseline %>%
#   left_join(., p_val_data_sdb) %>%
#   left_join(., conf_data_sdb)
# 
# sdb_baseline_out = sdb_baseline_out %>%
#   mutate_if(is.numeric, ~replace(., is.na(.), '')) %>%
#   mutate_if(is.character,~replace(., is.na(.), ''))
# 
# names(sdb_baseline_out) = c("Characteristics", paste0('All participants\n','n=', nrow(data)), paste0("PSQ > 0.33\n", "n=", sum(data$psq_binary == 1)), paste0("PSQ < 0.33\n", "n=", sum(data$psq_binary == 0)), "p-value", "CI")



sdb_endline_out = sdb_endline %>%
  left_join(., p_val_data_sdb_end) %>%
  left_join(., conf_data_sdb_end)

sdb_endline_out = sdb_endline_out %>%
  mutate_if(is.numeric, ~replace(., is.na(.), '')) %>%
  mutate_if(is.character,~replace(., is.na(.), ''))


names(sdb_endline_out) = c("Characteristics", paste0('All participants\n','n=', nrow(data)), paste0("PSQ > 0.33\n", "n=", sum(data$psq_b_binary == 1)), paste0("PSQ < 0.33\n", "n=", sum(data$psq_b_binary == 0)), "p-value", "CI")

# endline table

options("ReporteRs-fontsize" = 12);options("ReporteRs-default-font" = "Calibri")

table.endline <- sdb_endline_out %>% vanilla.table(.) %>% 
  setZebraStyle(., odd = '#eeeeee', even = 'white')
table.endline[,1] = parLeft()

doc = docx() %>%

  # Add spaces to center the cover page
  addParagraph(., c("\n")) %>% addParagraph(., c("\n")) %>% addParagraph(., c("\n")) %>%
  

addSection(., landscape = FALSE)%>% # Change Orientation of the page
addTitle(., value = 'TABLE 1. characteristics of all participants with and without SDB, 6-8 weeks post operation') %>%
addFlexTable(., flextable = table.endline)

writeDoc(doc, file = file.path(cache_dir, paste('endline characteristics.docx')))
knitr::kable(mtcars[1:10,], caption = "Characteristics of participants with and without SDB at 6-8 weeks post operation")
```

A total of `r nrow(filter(data, psq_b_binary == 1))` (`r round(nrow(filter(data, psq_b_binary == 1))/nrow(data)*100)`) childen and caregivers interviewed after treatment of SDB related symptoms showed signs of SDB (PSQ score >0.33). There were however no significant inter group differences in overal characteristics between chilldren with scores over 0.33 and those with scores lower than 0.33. More descriptive results are as presented in the table \@ref(tab:endline). <br>


```{r bar-plots-baseline, fig.cap= "Class separation of categorical dependent variables across the two levels of PSQ score"}

knitr::opts_chunk$set(chunk_option1 = TRUE)

data_viz = data %>%
  mutate(., asthma_diag = ifelse(asthma_diag == 0, 'yes', 'no'),
         allergic_rhinitis_diag = ifelse(allergic_rhinitis_diag == 1, 'yes', 'no'),
         sex = ifelse(sex == 1, 'male', 'female'))

data_viz = data_viz %>%
  mutate(., psq_binary = ifelse(psq_binary == 1, "PSQ > 0.33", "PSQ <= 0.33"))
library(gridExtra)
plot_bars <- function(df, catcols){
  options(repr.plot.width = 6, repr.plot.height = 5)
  temp0 <- df[df$psq_binary == "PSQ <= 0.33", ]
  temp1 <- df[df$psq_binary =="PSQ > 0.33", ]
  for (col in cat_cols){
    p1 <- ggplot(temp0, aes_string(col)) +
      geom_bar(fill = "#56B4E9", alpha = 0.9) +
      ggtitle(paste(gsub("_|diag"," ", col), '\nPSQ < 0.33')) + 
      coord_cartesian(clip = "off") +
      xlab("") +
    scale_y_continuous(
    name = "count",
    limits = c(0, 60),
    breaks = c(0, 4, 25,40),
    expand = c(0, 0)
    )+
    theme_minimal_hgrid(12, rel_small = 1)#+
    #   theme(
    #   #axis.ticks.length = grid::unit(0, "pt"),
    #   axis.ticks.x = element_blank(),
    #   axis.line = element_blank(),
    #   plot.margin = margin(3, 7, 3, 1.5)
    # )
    p2 <- ggplot(temp1, aes_string(col)) +
      geom_bar(fill = "#56B4E9", alpha = 0.9) +
      ggtitle(paste(gsub("_|diag"," ", col), '\nPSQ > 0.33')) +
       coord_cartesian(clip = "off") +
      xlab("") +
      scale_y_continuous(
    name = "count",
    limits = c(0, 60),
    breaks = c(0, 4, 25,40),
    expand = c(0, 0)
    )+
      theme_minimal_hgrid(12, rel_small = 1) #+
    #   theme(
    #   #axis.ticks.length = grid::unit(0, "pt"),
    #   axis.ticks.x = element_blank(),
    #   axis.line = element_blank(),
    #   plot.margin = margin(3, 7, 3, 1.5)
    # )
    grid.arrange(p1, p2, nrow = 1)
  }
}
cat_cols = c("age_cat", "asthma_diag","allergic_rhinitis_diag", "sex", "hospital", "`birth weight`", "psq_binary")
plot_bars(data_viz, cat_cols)
    # scale_y_continuous(expand = c(0, 0),
    #                    name = "median income (USD)",
    #                    breaks = c(0, 20000, 40000, 60000),
    #                    labels = c("$0", "$20,000", "$40,000", "$60,000")) +
    # xlab("age (years)") +
    # coord_cartesian(clip = "off") +
    # theme_dviz_hgrid() +
    # theme(
    #   #axis.ticks.length = grid::unit(0, "pt"),
    #   axis.ticks.x = element_blank(),
    #   axis.line = element_blank(),
    #   plot.margin = margin(3, 7, 3, 1.5)
    # )

```


To better understand class separation between the two PSQ score levels, we plotted bar graphs of age category, asthma, allergic rhinitis, birth weight, hospital status and of the PSQ score levels, figure \@ref(fig:bar-plot-baseline). Notably, there were some trend across the PSQ score levels; with age, asthma, allergic rhinitis, gender and hospital type categories having more numbers for scores of PSQ > 0. 33, when compared to a score of PSQ < 0.33. This observation indicates the possibility of these independent categorical explaining variability in PSQ scores. <br> 


```{r box-plot-baseline, fig.cap= "variation of numeric dependent variables across different levels of PSQ score"}
knitr::opts_chunk$set(chunk_option1 = TRUE)

plot_box <- function(df, cols, col_x = 'psq_binary'){
  options(repr.plot.width = 4, repr.plot.height = 3.5)
  for (col in cols){
    p <- ggplot(df, aes_string(col_x, col)) + 
        geom_boxplot(fill = 'grey90') +
         ggtitle(paste(gsub("_"," ", col))) + 
        coord_cartesian(clip = "off") +
        xlab("") +
        ylab("") +
        theme_minimal_hgrid(12, rel_small = 1)
    print(p)
     
  }
}

num_cols = c('height', 'wealth_quantile', "bmi")
plot_box(data_viz, num_cols)
```

We also visualized this class separation across numeric independent variables. There was clear class separation in the distribution of height across the two PSQ score levels. No particular trend was observed for bmi and wealth quantile as shown in Figure \@ref(fig:box-plot-baseline). <br>



A pie chart diagram of bmi category shows a missing class `obese` for PSQ scores > 0.33. This in a way limits modelling when some classes have cell counts of zero across variable of interest. <br>

```{r pie-plot-baseline, fig.cap= "variation of bmi categories for PSQ score > 0.33"}
knitr::opts_chunk$set(chunk_option1 = TRUE)

# radius of of the pie and radius for outside and inside labels
rpie = 1
rlabel_out = 1.05*rpie
rlabel_in = 0.6*rpie

bund_pie = data%>%
  filter(psq_binary == 1) %>%
  group_by(bmi_cat) %>%
   tally() %>%
  arrange(n) %>%
  mutate(end_angle = 2*pi*cumsum(n)/sum(n), # ending angle for each pie slice
         start_angle = lag(end_angle, default = 0), # starting angle for each pie slice
         mid_angle = 0.5*(start_angle + end_angle), # middle of each pie slice, for the text label
          # horizontal and vertical justifications depend on whether we're to the left/right
          # or top/bottom of the pie
         hjust = ifelse(mid_angle > pi, 1, 0),
         vjust = ifelse(mid_angle < pi/2 | mid_angle > 3*pi/2, 0, 1),
         colors = c("#E7D739", "#E64D93", "#B6494A","#4E4E4E")
         )

p1 = ggplot(bund_pie) +
  geom_arc_bar(
    aes(
      x0 = 0, y0 = 0, r0 = 0, r = rpie,
      start = start_angle, end = end_angle, fill = colors
    ),
    color = "white"
  ) +
  geom_text(
    aes(
      x = rlabel_in * sin(mid_angle),
      y = rlabel_in * cos(mid_angle),
      label = n
    ),
    size = 14/.pt,
    color = c("black", "white", "white", "white")
  ) +
  geom_text(
    aes(
      x = rlabel_out * sin(mid_angle),
      y = rlabel_out * cos(mid_angle),
      label = bmi_cat,
      hjust = hjust, vjust = vjust
    ),
    size = 10/.pt
  ) +
  scale_x_continuous(
    name = NULL,
    limits = c(-1.5, 1.4),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = NULL,
    limits = c(-1.05, 1.15),
    expand = c(0, 0)
  ) +
  scale_fill_identity() +
  coord_fixed() +
  theme_map()
bund_pie = data%>%
  filter(psq_binary == 0) %>%
  group_by(bmi_cat) %>%
   tally() %>%
  arrange(n) %>%
  mutate(end_angle = 2*pi*cumsum(n)/sum(n), # ending angle for each pie slice
         start_angle = lag(end_angle, default = 0), # starting angle for each pie slice
         mid_angle = 0.5*(start_angle + end_angle), # middle of each pie slice, for the text label
          # horizontal and vertical justifications depend on whether we're to the left/right
          # or top/bottom of the pie
         hjust = ifelse(mid_angle > pi, 1, 0),
         vjust = ifelse(mid_angle < pi/2 | mid_angle > 3*pi/2, 0, 1),
         colors = c("#E7D739", "#B6494A","#4E4E4E")
         )

p2 = ggplot(bund_pie) +
  geom_arc_bar(
    aes(
      x0 = 0, y0 = 0, r0 = 0, r = rpie,
      start = start_angle, end = end_angle, fill = colors
    ),
    color = "white"
  ) +
  geom_text(
    aes(
      x = rlabel_in * sin(mid_angle),
      y = rlabel_in * cos(mid_angle),
      label = n
    ),
    size = 14/.pt,
    color = c("black", "white", "white")
  ) +
  geom_text(
    aes(
      x = rlabel_out * sin(mid_angle),
      y = rlabel_out * cos(mid_angle),
      label = bmi_cat,
      hjust = hjust, vjust = vjust
    ),
    size = 10/.pt
  ) +
  scale_x_continuous(
    name = NULL,
    limits = c(-1.5, 1.4),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = NULL,
    limits = c(-1.05, 1.15),
    expand = c(0, 0)
  ) +
  scale_fill_identity() +
  coord_fixed() +
  theme_map()

p1
```


```{r pie2-plot-baseline, fig.cap= "variation of bmi categories for PSQ score <= 0.33"}
knitr::opts_chunk$set(chunk_option1 = TRUE)
p2
```




```{r logit, fig.cap= "Estimates of logistic regression of PSQ score before operation"}

knitr::opts_chunk$set(chunk_option1 = TRUE)

Data = data
# data = Data


# data$wet_bed = as.factor(data$wet_bed)

# cols = names(data[,44:96])
# data[,cols] = sapply(data[, cols], as.character)
# data$allergic_rhinitis_diag = as.numeric(data$allergic_rhinitis_diag) %>%
#   as.character(.)

# data$osa_binary = as.character(data$osa_binary)
# 
# data$asthma_diag = as.character(data$asthma_diag)
# data$sex = as.character(data$sex)
# data$psq_binary = as.factor(data$psq_binary)

# data = data[,-c(108:114)]
# data = data[,-c(109:113)]

# data = data %>%
#   dplyr::select(-c(shake_breath_b, apnes_b))
# 
# data_reg = data %>%
#   dplyr::select(-c(sub_id, sub_id_b))


# split

# set.seed(123)
# 
# train_test_split = rsample::initial_split(
#   data = data_reg,
#   prop = 0.7,
#   strata = "psq_binary"
# )
# 
# # train_test_split
# 
# train_tbl = train_test_split %>% training()
# test_tbl = train_test_split %>% testing()
# 
# #  pre process
# 
# recipe_simple <- function(dataset) {
#   recipe(psq_binary ~ ., data = dataset) %>%
#     step_string2factor(all_nominal(), -all_outcomes()) %>%
#     prep(data = dataset)
# }
# 
# recipe_prepped = recipe_simple(dataset = train_tbl)
# 
# train_baked <- bake(recipe_prepped, new_data = train_tbl)
# test_baked  <- bake(recipe_prepped, new_data = test_tbl)

# fit model

# logistic_glm <-
#   logistic_reg(mode = "classification") %>%
#   set_engine("glm") %>%
#   fit(psq_binary ~ age_cat+hospital+bmi_cat+allergic_rhinitis_diag+asthma_diag+height+`birth weight`, data = train_baked)
# predictions_glm <- logistic_glm %>%
#   predict(new_data = test_baked) %>%
#   bind_cols(test_baked %>% dplyr::select(psq_binary))
# head(predictions_glm)

# predictions_glm %>%
#   conf_mat(psq_binary, .pred_class) %>%
#   pluck(1) %>%
#   as_tibble() %>%
#   ggplot(aes(Prediction, Truth, alpha = n)) +
#   geom_tile(show.legend = FALSE) +
#   geom_text(aes(label = n), colour = "white", alpha = 1, size = 8)

# predictions_glm %>%
#   metrics(psq_binary, .pred_class) %>%
#   dplyr::select(-.estimator) %>%
#   filter(.metric == "accuracy") 
# 
# tibble(
#   "precision" = 
#      precision(predictions_glm, psq_binary, .pred_class) %>%
#      dplyr::select(.estimate),
#   "recall" = 
#      recall(predictions_glm, psq_binary, .pred_class) %>%
#      dplyr::select(.estimate)
# ) %>%
#   unnest() -> precision
#   # kable() 
# 
# t = tidy(logistic_glm)
# kable(t, caption = "Logistic regression coefficients, SD and p-value")
# # lapply(data, unique)

col = c("osa_binary", "wet_bed", "allergic_rhinitis_diag", "asthma_diag", "antiacids", "smoking")
for(c in col){
data[, c] = sapply(data[,c], function(x){
  x = ifelse(x == 0, 'No', 'Yes')
  x = factor(x, levels = c('No', 'Yes'))
})
}

logistic_glm <-glm(psq_binary ~ allergic_rhinitis_diag+asthma_diag+height+`birth weight`+age_cat+sex+hospital+bmi_cat, data = data)

# summary(logistic_glm)
estimates = tidy(logistic_glm)
ci = confint.default(logistic_glm)
interval = tidy(ci)
names(interval) = c("term", "`2.5 %`", "`97.5 %`")
estimates = estimates %>%
  left_join(interval)
kable(estimates)
# data %>% dplyr::filter(., psq_b_binary == 1) %>% count(allergic_rhinitis_diag)
```


## Logistic regression results

We further performed a logistic regression to explore predictors of SDB scores > 0.33. Overal, having allergic rhinitis contributed a significant change in log odds of SDB by (`r paste('0.2322', '0.10244', sep = "\u00B1" )`), p-value = 0.0266. Accessing treatment of SDB in a public hospital also had an effect of 0.3032 on log odds of having SDB when compared to going to a private hospital. Inter group difference in other categorical factors like bmi category, age, birth weight category, gender were not significant as shown in table \@ref(tab:logit) <br>. This was a very simple logistic regression model with minimal paremtric tuning, but which can be improved if more data indicative of SDB were available. <br>


```{r glm-test, echo=FALSE}
knitr::opts_chunk$set(chunk_option1 = TRUE)
library(aod)

chi_glm = wald.test(b = coef(logistic_glm), Sigma = vcov(logistic_glm), Terms = 9:11)
a = chi_glm$result[[1]][3] # p-value

```

We tested for the overall effect of bmi category and it wasn't statistically significant (p = 0.4322). One explanation for this is the existence of null cells across PSQ score levels. <br> 

```{r odds-ratio, echo=FALSE, fig.cap="Odds of the logistic regression and CIs"}
knitr::opts_chunk$set(chunk_option1 = TRUE)
odds = exp(cbind(OR = coef(logistic_glm), confint(logistic_glm))) %>% tidy()

names(odds) = c("term", "OR", "`2.5 %`", "`97.5 %`")

kable(odds)


```


An interpretation of the odds indicate that having allergic rhinitis increased the odds of a having PSQ score > 0.33 by a factor of 1.26. An admission into a public hospital was also associated by an increase in the odds of having SDB by a factor of 1.35. A similara infrence can be drawn for remainder of the factors shown in table \@ref(tab:odds-ratio). <br>


## General estimating equation model


```{r gee, fig.cap= "Estimates of logistic regression of PSQ score after operation"}

knitr::opts_chunk$set(chunk_option1 = TRUE)
library(geepack)
library(glmnet)

data = filter(data, !is.na(wet_bed))
train_x <- model.matrix(psq_b_binary ~ age+osa_binary+allergic_rhinitis_diag+asthma_diag+`birth weight`+antiacids+smoking+bmi+wet_bed, data)[,-1]
a = as.numeric(row.names(train_x))
train = filter(data, as.numeric(row.names(data)) %in% a)
# create training and testing response vectors
train_y <- train$psq_b_binary
ridge <- glmnet(x = train_x, y = train_y, alpha = 0.2, family = "binomial")
print(ridge)
plot(ridge)
a = coef(ridge,s=0.1) %>% yardstick::tidy()


kable(a[-1,])
# ridge$lambda.min
# cvfit = cv.glmnet(x = train_x, y = train_y, alpha = 0.2, family = "binomial")
# plot(cvfit)
# cvfit$lambda.min
# coef(cvfit, s = "lambda.min")
psq_gee2 <- gee(psq_b_binary  ~ age+osa_binary+allergic_rhinitis_diag+asthma_diag+`birth weight`+antiacids+bmi+wet_bed, data = data, family = "binomial", id = sub_id, corstr = "exchangeable", scale.fix = TRUE,scale.value = 1)


cols = col = c("osa_binary", "wet_bed", "allergic_rhinitis_diag", "asthma_diag", "antiacids", "smoking")


# data %>% dplyr::filter(., psq_b_binary == 1) %>% count(wet_bed)

# crude odds

yyyy = list()
dep = c("allergic_rhinitis_diag", "asthma_diag", "antiacids","birth weight", "age", "osa_binary", "bmi", "wet_bed")
for(x in dep) {
  model = gee(psq_b_binary ~ data[,x], data = data, family = "binomial", id = sub_id, corstr = "exchangeable", scale.fix = TRUE,scale.value = 1) # two angle brackets
  est = summary(model)
  coef = yardstick::tidy(est$coefficients)
  yyyy[[x]] = list(variable = x, coef =coef$Estimate[2], `crude odds` = exp(coef$Estimate)[2])
  
}
yy = yyyy %>% dplyr::bind_rows()

estimates = summary(psq_gee2)
coef = yardstick::tidy(estimates$coefficients)
coef = coef %>%  dplyr::select(c(.rownames,Estimate,Robust.S.E.)) %>% `colnames<-`(c('variable', 'coef', 'se'))
coef$adjusted_odds = exp(coef$coef)
coef$lower_ci = round(exp(coef$coef +
c(-1) * coef$se * qnorm(0.975)),4)
coef$upper_ci =round(exp(coef$coef +
c(1) * coef$se * qnorm(0.975)),4)

skimr::kable(yy)

```

```{r coef, fig.cap= "Estimates of gee of PSQ score after operation"}

skimr::kable(coef)
```








