---
output: officedown::rdocx_document
---

```{r setup, include=FALSE}
library(officedown)
library(ggplot2)
```

## Figure

```{r fig.cap="economics plot", fig.id = "tsplot", fig.cap.style = "Image Caption"}
ggplot(economics, aes(date, unemploy / pop)) + 
  geom_line() + 
  theme_minimal()
```

## Tables

### economics

```{r tab.cap="economics table", tab.id = "mytab", tab.cap.style = "Table Caption"}
head(economics)
```

### mtcars

```{r tab.cap="mtcars table", tab.id = "mtcars", tab.cap.style = "Table Caption"}
head(mtcars)
```

## Cross-referencing

Figure and table have auto-numbered captions that can be cross referenced: 

* This is a linked reference to a figure \@ref(fig:tsplot), its number is computed by Word 
and it's linked to the corresponding graphic when clicking on it.
* This is a linked reference to a table \@ref(tab:mytab), its number is computed by Word 
and it's linked to the corresponding table when clicking on it. This is a linked reference to 
a table \@ref(tab:mtcars).

