---
title: "Table captions linked to section number"
output: officedown::rdocx_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(officedown)
library(officer)
```

## Tables

### Table example

The code below define the set of runs that will compose the auto-numbered sequence in the caption:

```{r}
runs <- list(
  run_word_field("STYLEREF 3 \\s"),
  ftext(" - "),
  run_autonum(seq_id = "tab", pre_label = "", post_label = "")
)
```

And now we can add this sequence into a custom caption:

::: {custom-style="Table Caption"}



Figure `r run_bookmark("airquality", runs)`: truc machin chose



:::

```{r}
head(airquality)
```

This is a reference to bookmark `airquality`: `r run_reference(id = "airquality")`

```{r}
runs_0 <- list(ftext("Figure :"))
runs_p <- list(ftext(": mtcars dataset"))
runs <- append(runs_0, runs)
runs <- append(runs, runs_p)
```

::: {custom-style="Table Caption"}



`r run_bookmark("mtcars", runs)`



:::

```{r}
head(mtcars)
```

This is a reference to bookmark `mtcars`: `r run_reference(id = "mtcars")`