---
title: "Chemoprevention PhD Studentship"
subtitle: "Current work: "
author: "Basil okola"
date: "`r format(Sys.Date(),'%m %d, %Y')`"
header-includes: 
  - \usepackage{amsmath}
output:
  xaringan::moon_reader:
    css: ["default", "css/my-theme.css", "css/my-fonts.css"]
    seal: false
    lib_dir: libs
    nature:
      ratio: '4:3'
      highlightStyle: dracula
      highlightLines: true
      countIncrementalSlides: false
toc: yes
---

```{r, child="_setup.Rmd"}

```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE,
        eval = TRUE)
library(tidyverse)
library(testthat)
library(gt)
library(devtools)
library(usethis)
library(here)
library(readxl)
```

```{r metathis, echo=FALSE}
library(metathis)
meta() %>%
  meta_name("github-repo" = "jthomasmock/pkg-building") %>% 
  meta_social(
    title = "Package Building with devtools and usethis",
    description = paste(
      "Package building doesn't have to be scary!",
      "The tidyverse team made it easy to get started",
      "with RStudio, {devtools}, & {usethis}."
    ),
    url = "https://colorado.rstudio.com/rsc/pkg-building/",
    image = "https://raw.githubusercontent.com/jthomasmock/pkg-building/master/images/title-card.png",
    image_alt = paste(
      "Title slide of Package Building with devtools and usethis.", 
      "A slide deck covering the basics of writing functions and developing packages.",
      "It also has an image of a orange rubik cube with a blue heart in the middle."
    ),
    og_type = "website",
    og_author = "Tom Mock",
    twitter_card_type = "summary_large_image",
    twitter_creator = "@thomas_mock",
    twitter_site = "@thomas_mock"
  )
```

class: title-slide, left, top

# `r rmarkdown::metadata$title`

## `r rmarkdown::metadata$subtitle`

### `r rmarkdown::metadata$author`

### `r format(Sys.Date(), '%b %d, %Y')`

<br>

---

# Introduction

- Reported cases of COVID-19 stand at about <b>758M</b> with deaths about <b>6.9M</b>

- A number of coping measures including restricted movement, lockdowns, closure of   schools, social joints were implemented

- While the spread is now under control thanks to vaccines, the effects live on

- There are reported cases of long COVID

- Growing interest in understanding the mental burden of the disease

- Coping measures + losing close family members,  contracting covid (could) have greater impact on mental health of survivors

---

# Data

- Omtanke2020 (since September 2020) study hosted by KI, Sweden

- A self-reported longitudinal data on:

  - socio-demographics: age, sex, relationship status, general lifestyle
  mental and physical health, COVID-19 infection and impact

  - Depression via 9-item questions Patient Health Questionnaire (PHQ-9) on a four-point Likert scale 
<!-- (0 = not at all, 1 = somedays, 2 = more than half of the days, 3 = nearly every day). -->
  - Anxiety via 7-item questions General Anxiety Disorder (GAD-7) on a four-point Likert scale

  - COVID-19-related distress using COVID-19 primary care PTSD screening tool
---
# Methods

- Poisson regression used to assess the relative risk of demonstrating high level symptoms of depression,

# Preliminary results

