---
title: "Mortgage Classification"
author: "bokola"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    highlight: tango
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 3
---

# **Background**
The objective is Predicting Mortgage Approvals From Government Data, taking into consideration of **demographics, location, property type, lender**, and other factors are related to whether a mortgage application was accepted or denied,applying relevant skills to build a model for predicting **acceptance** of loan application across the United States. There are 23 variables (21 possible features and 1 binary label(`accepted`)) and 1 arbitrary identifier(`row_id). 

```{r packages, warning=FALSE,results='hide', echo=FALSE, message=FALSE}
rm(list = ls(all = T))
#setwd("C:\\Users\\bokola\\Google Drive\\Data")
ipk <- function(pkg){
  new.pkg <- list.of.pkg[!(list.of.pkg %in% installed.packages()[, "Package"])]
  if(length(new.pkg)) install.packages(new.pkg, dependencies = T)
  sapply(pkg, require, character.only = T)
  
}
list.of.pkg <- c("plyr", "tidyverse", "repr", "magrittr","kknn", "gridExtra", "GGally", "hexbin", "gridExtra")
ipk(list.of.pkg)
options(repr.plot.width = 4, repr.plot.height = 3.5)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#knitr::opts_knit$set(root.dir = "C:\\Users\\bokola\\Google Drive\\Data")
hom.dir = ifelse(Sys.info()["sysname"] == "Windows", Sys.getenv("USERPROFILE"), Sys.getenv("HOME"))
project.path = path.expand(file.path(hom.dir
                                     ,"Analysis"
                                     ,"capstone")) %>% gsub("\\\\", "/", .)

data.path = path.expand(file.path(project.path
                                  ,"Data"))
scripts.path = path.expand(file.path(project.path
                                     ,"src"
                                     ,"r"))
report.path = path.expand(file.path(project.path
                                    ,"cache"
                                    ,"doc"))
if(!file.exists(project.path)){
  if(dir.create(project.path, recursive = T))
    stop("The project directory \"",
         project.path, "\"has been created!\nPlease fill it with relevant files and folders!")
  else
    stop("The project directory\"",
         project.path,
         "\"could not be created!")
}
if(!file.exists(data.path)){
  if(dir.create(data.path, recursive = T))
    stop("The project directory \"",
         data.path, "\"has been created!\nPlease fill it with relevant files and folders!")
  else
    stop("The project directory\"",
         data.path,
         "\"could not be created!")
}
if(!file.exists(report.path)){
  if(dir.create(report.path, recursive = T))
    stop("The project directory \"",
         report.path, "\"has been created!\nPlease fill it with relevant files and folders!")
  else
    stop("The project directory\"",
         report.path,
         "\"could not be created!")
}
if(!file.exists(scripts.path)){
  if(dir.create(scripts.path, recursive = T))
    stop("The project directory \"",
         scripts.path, "\"has been created!\nPlease fill it with relevant files and folders!")
  else
    stop("The project directory\"",
         scripts.path,
         "\"could not be created!")
}
```

Data import

```{r Data, warning=FALSE,results='hide', echo=FALSE, message=FALSE}
train_labels = read.csv(path.expand(file.path(data.path, "train_labels.csv")), header = T)
train_values = read.csv(path.expand(file.path(data.path, "train_values.csv")), header = T)
test_values = read.csv(path.expand(file.path(data.path, "test_values.csv")), header = T) 
train = merge(train_labels, train_values)

# dim(train) 
# head(train)

#Transformation
cols = c('msa_md', 'state_code','county_code','lender','loan_type','property_type', 'loan_purpose', 'occupancy','preapproval','applicant_ethnicity','applicant_race', 'applicant_sex')
train[,cols] = lapply(train[,cols], as.character)
test_values[,cols] = lapply(train[,cols], as.character)
#1. loan type
loan_type= c('1' = 'Conventional', '2' = 'FHA-insured ', '3' = 'VA-guaranteed', "4" = 'FSA/RHS')
property_type = c("1" = "One to four-family", "2" = "Manufactured housing", "3" = "Multifamily")
loan_purpose = c("1" = "Home purchase", "2" = "Home improvement", "3" = "Refinancing")
occupancy = c("1" = "Owner-occupied", "2" = "Not owner-occupied", "3" = "Not applicable")
preapproval = c("1" = "requested", "2" = "not requeste", "3" = "not applicable")
applicant_ethnicity = c("1" = "Hispanic or Latino", "2" = "Not Hispanic or Latino", "3" = "no information", "4" = "not applicable")#, "5" = "no co-applicant"
applicant_race = c("1" = "American Indian/Alaska Native", "2" = "Asian", "3" = "Black/African American", "4" =  "Native Hawaiian", "5" = "White", "6" = "no information", "7" = "not applicable")#, "8" = "no co-applicant"
applicant_sex = c("1" = "Male", "2" = "Female", "3" = "no information", "4" = "not applicable")

# out = rep('i', length.out = nrow(train))
# i=1
# for (x in train[,'loan_type']) {
#   out[i] = loan_type[[x]]
#   i=i+1
# }
# train[,"loan_type"] = out

for(var in c("loan_type","property_type","loan_purpose","occupancy", "preapproval", "applicant_ethnicity", "applicant_race", "applicant_sex")){
  train[,var]=sapply(train[,var],function(x) get(var)[[x]],simplify = T,USE.NAMES = F)
  #test_values[,var]=sapply(test_values[,var],function(x) get(var)[[x]],simplify = T,USE.NAMES = F)
}
a = list(train, test_values)

# v = function(y){
#   for(v in c("loan_type","property_type","loan_purpose","occupancy", "preapproval", "applicant_ethnicity", "applicant_race", "applicant_sex")){
#   #train[,var]=sapply(train[,var],function(x) get(var)[[x]],simplify = T,USE.NAMES = F)
#   y[,v]=sapply(y[,v],function(x) get(var)[[x]],simplify = T,USE.NAMES = F)
#   }
# }

for(v in c("loan_type","property_type","loan_purpose","occupancy", "preapproval", "applicant_ethnicity", "applicant_race", "applicant_sex")){
  #train[,var]=sapply(train[,var],function(x) get(v)[[x]],simplify = T,USE.NAMES = F)
  test_values[,v]=sapply(test_values[,v],function(x) get(v)[[x]],simplify = T,USE.NAMES = F)
}
#train[,cols] = lapply(train[,cols], as.factor)
#test_values[,cols] = lapply(train[,cols], as.factor)

#,"loan_purpose", "occupancy", "preapproval","applicant_ethnicity","applicant_race", "applicant_sex"
#ss=sapply(train[,var],function(x) get(var)[[x]],simplify = T,USE.NAMES = F)

# for(col in names(train)) {
#   if(is.character(train[, col])) {
#     count <- sum(ifelse(train[, col] == '-1'|is.na(train[,col]), 1, 0))
#     cat(paste(col, as.character(count), '\n'))
#   }
# }

```
Summary statistics
```{r summaries, warning=FALSE,results='hide', echo=FALSE, message=FALSE}
train %>% summarise(min = min(loan_amount, na.rm = T), max = max(loan_amount, na.rm = T), median = median(loan_amount, na.rm = T), mean = mean(loan_amount, na.rm = T), sd = sd(loan_amount, na.rm = T))
train %>% aggregate(accepted ~ applicant_ethnicity, ., mean)
train %>% aggregate(accepted ~ applicant_sex, ., mean)
r = train %>% subset(., state_code == "48" & county_code != "-1") %>%  aggregate(accepted ~ county_code, ., mean)
range(r$accepted)
train %>% subset(., state_code %in% c("2","4")) %>% aggregate(accepted ~ state_code+loan_type, ., mean) %>% arrange(loan_type, accepted)

plot_hist <- function(df, col = 'loan_amount', bins = 4){
  #options(repr.plot.width = 4, repr.plot.height = 3.5)
  #bw <- (max(df[, col]) - min(df[, col])) / (bins+1)
  p <- ggplot(df, aes_string(col)) +
    #geom_histogram(binwidth = bw) + scale_x_continuous(breaks = seq(0,1200, 200))#, aes(y = ..density..),alpha = 0.5) #+
  geom_histogram()+ scale_x_log10(breaks = seq(0,1200, 100))#(breaks = seq(0,1200, 200))
    #geom_density(aes(y = ..density..), color = 'blue') +
    #geom_rug()
  print(p)
}
plot_hist(train)
df = subset(train, loan_amount <=1200)
df = df %>% 
  mutate
(p = ggplot(subset(train, state_code == "45"), aes(x = applicant_income,y = loan_amount)) + geom_point() + geom_smooth())
  
```
# **Data Preparation and Visualization**
Data preparation is an important step. We first **Scale** the numeric values of the features to ensure they have a similar range of values to avoid features with larger numeric values dominating model training. `Zscore` normalization is used. We then recode categorical variable to factors. We then  visualize the separation of values between different levels of the label - `accepted(0,1)`
1. **Class separation by numeric variables**
We explore the distinction of the distribution of values between the two levels of the label.
1. Boxplot

```{r Data transform, warning=FALSE, echo=FALSE, message=FALSE}
 numeric_vars<-sapply(names(train),function(x) ifelse(is.numeric(train[,x]) & x !="row_id",x,NA))
 numeric_vars<-na.omit(numeric_vars)

 cat_vars<-sapply(names(train),function(x) ifelse(is.character(train[,x]),x,NA))
 cat_vars<-na.omit(cat_vars)
numeric_vars = function(x) {
  x = sapply(names(x), function(x) ifelse(is.numeric(train[,x]) & x !="row_id" & x != "accepted",x,NA))
  x = na.omit(x)
  
} 

cat_vars = function(x) {
  x = sapply(names(x), function(x) ifelse(is.factor(train[,x]),x,NA))
  x = na.omit(x)
  
} 

#train[, numeric_vars(train)] = sapply(train[, numeric_vars(train)], scale)
#test_values[, numeric_vars(test_values)] = sapply(test_values[, numeric_vars(test_values)], scale)
#ss=apply(train[, sapply(train, is.numeric)],2, scale)
train$accepted = as.factor(train$accepted)
plot_box <- function(df, cols, col_x = 'accepted'){
  options(repr.plot.width = 4, repr.plot.height = 3.5)
  for (col in cols){
    p <- ggplot(df, aes_string(col_x, col)) + 
        geom_boxplot() +
        ggtitle(paste('Box plot of', col, '\n vs.', col_x))
    print(p)
     
  }
}

num_vars = c("loan_amount", "applicant_income", "population", "minority_population_pct", "ffiecmedian_family_income", "tract_to_msa_md_income_pct", "number_of_owner.occupied_units", "number_of_1_to_4_family_units")
plot_box(train, num_vars)
```
From the plots, three variables - loan_amount, minority_population_pct and tract_to_msa_md_income_pct show clear separation of values between the two levels. Next we explore this separation by categorical variables.
2. Bar graphs
```{r Data transform, warning=FALSE, echo=FALSE, message=FALSE}
plot_bars = function(df, catcols){
  options(repr.plot.width = 6, repr.plot.height = 5)
  temp0 = df[df$accepted == 0,]
  temp1 = df[df$accepted == 1,]
  for(col in cat_cols){
    p1 = ggplot(temp0, aes_string(col)) + 
      geom_bar() + 
      ggtitle(paste('Bar plot of \n', col, '\n for rejected loan applications')) + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1))
    p2 = ggplot(temp1, aes_string(col)) + 
      geom_bar() + 
      ggtitle(paste('Bar plot of \n', col, '\n for approved loan applications')) + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1))
    grid.arrange(p1, p2, nrow = 1)
  }
}
cat_cols = c("loan_type","property_type","loan_purpose","occupancy", "preapproval", "applicant_ethnicity", "applicant_race", "applicant_sex")
plot_bars(train, cat_cols)
```
We observe the following:
1.	Some features such as property_type, loan_purpose, applicant_ethnicity, applicant_race have an almost significant different distribution of categories between the label categories.
2.	Others features such as preapproval, applicant_sex, occupancy show small differences, but these differences are unlikely to be significant.

Next we do feature engineering in readiness for modelling.




