---
title: "Mortgage Classification using CatBoostclassifier"
author: "bokola"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  md_document:
    #variant: markdown_github
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    highlight: tango
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 3
---

# **Executive Summary**

The objective is Predicting Mortgage Approvals From Government Data, taking into consideration the **demographics, location, property type, lender**, and other factors to predict whether a mortgage application was accepted or denied. It is a classification problem - A classifier is a machine learning model that separates the **label** into categories or **classes**. In other words, classification models are **supervised** machine learning models which predict a categorical label. We apply relevant skills to classify the label **`accepted`** - whether a mortgage application was accepted or declined using data obtained across the United States.

We used a CatBoostclassifier model with 1000 iterations with 8 cross-validations with an accuracy of 0.72.


# **Methodology**

The section outlines data manipulation procedures, visualizations and model selection.

## **Description of Data**

Data consisted of 23 variables (21 possible features and 1 binary label(`accepted`) and 1 arbitrary identifier(`row_id`). Both training and test sets spanned 500000 records


```{r packages, warning=FALSE,results='hide', echo=FALSE, message=FALSE}
rm(list = ls(all = T))
#setwd("C:\\Users\\bokola\\Google Drive\\Data")
ipk <- function(pkg){
  new.pkg <- list.of.pkg[!(list.of.pkg %in% installed.packages()[, "Package"])]
  if(length(new.pkg)) install.packages(new.pkg, dependencies = T, repos = "http://cran.us.r-project.org")
  sapply(pkg, require, character.only = T)
  
}
list.of.pkg <- c("RCurl", "plyr", "tidyverse", "repr", "magrittr","kknn", "gridExtra", "GGally", "hexbin", "gridExtra", "caret", "ROCR", "pROC", "mice", "glmnet", "recipes", "h2o", "reticulate")
ipk(list.of.pkg)
options(repr.plot.width = 4, repr.plot.height = 3.5)

# h2o.no_progress()
# # launch h2o
# h2o.init()
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
use_virtualenv("r-reticulate")
#knitr::opts_knit$set(root.dir = "C:\\Users\\bokola\\Google Drive\\Data")
```

```{r paths, include=F}

hom.dir = ifelse(Sys.info()["sysname"] == "Windows", Sys.getenv("USERPROFILE"), Sys.getenv("HOME"))
project.path = path.expand(file.path(hom.dir
                                     ,"Analysis"
                                     ,"capstone")) %>% gsub("\\\\", "/", .)

data.path = path.expand(file.path(project.path
                                  ,"Data"))
scripts.path = path.expand(file.path(project.path
                                     ,"src"
                                     ,"r"))
report.path = path.expand(file.path(project.path
                                    ,"cache"
                                    ,"doc"))
if(!file.exists(project.path)){
  if(dir.create(project.path, recursive = T))
    stop("The project directory \"",
         project.path, "\"has been created!\nPlease fill it with relevant files and folders!")
  else
    stop("The project directory\"",
         project.path,
         "\"could not be created!")
}
if(!file.exists(data.path)){
  if(dir.create(data.path, recursive = T))
    stop("The project directory \"",
         data.path, "\"has been created!\nPlease fill it with relevant files and folders!")
  else
    stop("The project directory\"",
         data.path,
         "\"could not be created!")
}
if(!file.exists(report.path)){
  if(dir.create(report.path, recursive = T))
    stop("The project directory \"",
         report.path, "\"has been created!\nPlease fill it with relevant files and folders!")
  else
    stop("The project directory\"",
         report.path,
         "\"could not be created!")
}
if(!file.exists(scripts.path)){
  if(dir.create(scripts.path, recursive = T))
    stop("The project directory \"",
         scripts.path, "\"has been created!\nPlease fill it with relevant files and folders!")
  else
    stop("The project directory\"",
         scripts.path,
         "\"could not be created!")
}
```




```{r Data, warning=FALSE,results='hide', echo=FALSE, message=FALSE}
train_labels = read.csv("https://www.dropbox.com/s/qvq3hg3mnlowxql/train_labels.csv?raw=1", header = T)
train_values = read.csv("https://www.dropbox.com/s/3l8eyc51iupjglf/train_values.csv?raw=1", header = T)
test_values = read.csv("https://www.dropbox.com/s/2heyrfbq3cxvsdb/test_values.csv?raw=1", header = T) 
train = merge(train_labels, train_values)

# distribution of NAs across variables
is_na = function(x) {
  for (col in names(x)) {
    #count = sum(which(x[,col] == "" | is.na(x[,col])))
    #count = sum(which(is.na(x[,col])))
    count = sum(is.na(x[,col]))
    cat(paste(col, ":", as.character(count), '\n'))
    
  }
}
is_na(train)
missing = c("applicant_income", "population", "minority_population_pct", "ffiecmedian_family_income", "tract_to_msa_md_income_pct", "number_of_owner.occupied_units", "number_of_1_to_4_family_units")
# train[, missing] = ifelse(is.na(train[,missing]), median(train[,missing]), train[, missing])
# train_fill = train[,missing]
# mice(data = train_fill, m = 5, method = "pmm", maxit = 5, seed = 500)
# train %<>% mutate_at(vars(applicant_income, tract_to_msa_md_income_pct), ~ifelse(is.na(.x), median(.x, na.rm = T), .x))
# test_values%<>% mutate_at(vars(applicant_income, tract_to_msa_md_income_pct), ~ifelse(is.na(.x), median(.x, na.rm = T), .x))
# dim(train) 
# head(train)

#Transformation
cols = c('msa_md', 'state_code','county_code','lender','loan_type','property_type', 'loan_purpose', 'occupancy','preapproval','applicant_ethnicity','applicant_race', 'applicant_sex')
train[,cols] = lapply(train[,cols], as.character)
test_values[,cols] = lapply(train[,cols], as.character)
#1. loan type
loan_type= c('1' = 'Conventional', '2' = 'FHA-insured ', '3' = 'VA-guaranteed', "4" = 'FSA/RHS')
property_type = c("1" = "One to four-family", "2" = "Manufactured housing", "3" = "Multifamily")
loan_purpose = c("1" = "Home purchase", "2" = "Home improvement", "3" = "Refinancing")
occupancy = c("1" = "Owner-occupied", "2" = "Not owner-occupied", "3" = "Not applicable")
preapproval = c("1" = "requested", "2" = "not requeste", "3" = "not applicable")
applicant_ethnicity = c("1" = "Hispanic or Latino", "2" = "Not Hispanic or Latino", "3" = "no information", "4" = "not applicable")#, "5" = "no co-applicant"
applicant_race = c("1" = "American Indian/Alaska Native", "2" = "Asian", "3" = "Black/African American", "4" =  "Native Hawaiian", "5" = "White", "6" = "no information", "7" = "not applicable")#, "8" = "no co-applicant"
applicant_sex = c("1" = "Male", "2" = "Female", "3" = "no information", "4" = "not applicable")

# out = rep('i', length.out = nrow(train))
# i=1
# for (x in train[,'loan_type']) {
#   out[i] = loan_type[[x]]
#   i=i+1
# }
# train[,"loan_type"] = out

for(var in c("loan_type","property_type","loan_purpose","occupancy", "preapproval", "applicant_ethnicity", "applicant_race", "applicant_sex")){
  train[,var]=sapply(train[,var],function(x) get(var)[[x]],simplify = T,USE.NAMES = F)
  #test_values[,var]=sapply(test_values[,var],function(x) get(var)[[x]],simplify = T,USE.NAMES = F)
}
a = list(train, test_values)

# v = function(y){
#   for(v in c("loan_type","property_type","loan_purpose","occupancy", "preapproval", "applicant_ethnicity", "applicant_race", "applicant_sex")){
#   #train[,var]=sapply(train[,var],function(x) get(var)[[x]],simplify = T,USE.NAMES = F)
#   y[,v]=sapply(y[,v],function(x) get(var)[[x]],simplify = T,USE.NAMES = F)
#   }
# }

for(v in c("loan_type","property_type","loan_purpose","occupancy", "preapproval", "applicant_ethnicity", "applicant_race", "applicant_sex")){
  #train[,var]=sapply(train[,var],function(x) get(v)[[x]],simplify = T,USE.NAMES = F)
  test_values[,v]=sapply(test_values[,v],function(x) get(v)[[x]],simplify = T,USE.NAMES = F)
}
#train[,cols] = lapply(train[,cols], as.factor)
#test_values[,cols] = lapply(train[,cols], as.factor)

#,"loan_purpose", "occupancy", "preapproval","applicant_ethnicity","applicant_race", "applicant_sex"
#ss=sapply(train[,var],function(x) get(var)[[x]],simplify = T,USE.NAMES = F)

# for(col in names(train)) {
#   if(is.character(train[, col])) {
#     count <- sum(ifelse(train[, col] == '-1'|is.na(train[,col]), 1, 0))
#     cat(paste(col, as.character(count), '\n'))
#   }
# }

```


There are a number of variables in the train and test sets with missing values: applicant_income - 39948, 
population - 22465, minority_population_pct - 22466, ffiecmedian_family_income - 22440, tract_to_msa_md_income_pct - 22514, number_of_owner.occupied_units - 22565, number_of_1_to_4_family_units - 22530. The missing entries were recoded to -999 for numeric variables.



```{r NAs, warning=FALSE,results='hide', echo=FALSE, message=FALSE}
p = train %>% 
  is.na() %>%
  reshape2::melt() %>%
  ggplot(aes(Var2, Var1, fill=value)) + 
    geom_raster() + 
    coord_flip() +
    scale_y_continuous(NULL, expand = c(0, 0)) +
    scale_fill_grey(name = "", labels = c("Present", "Missing")) +
    xlab("Observation") +
    theme(axis.text.y  = element_text(size = 4))
# vis_miss(train, cluster = TRUE)

```



```{r summaries, warning=FALSE,results='hide', echo=FALSE, message=FALSE}
train %>% summarise(min = min(loan_amount, na.rm = T), max = max(loan_amount, na.rm = T), median = median(loan_amount, na.rm = T), mean = mean(loan_amount, na.rm = T), sd = sd(loan_amount, na.rm = T))
train %>% aggregate(accepted ~ applicant_ethnicity, ., mean)
train %>% aggregate(accepted ~ applicant_sex, ., mean)
r = train %>% subset(., state_code == "48" & county_code != "-1") %>%  aggregate(accepted ~ county_code, ., mean)
range(r$accepted)
train %>% subset(., state_code %in% c("2","4")) %>% aggregate(accepted ~ state_code+loan_type, ., mean) %>% arrange(loan_type, accepted)

# plot_hist <- function(df, col = 'loan_amount', bins = 4){
#   #options(repr.plot.width = 4, repr.plot.height = 3.5)
#   #bw <- (max(df[, col]) - min(df[, col])) / (bins+1)
#   p <- ggplot(df, aes_string(col)) +
#     #geom_histogram(binwidth = bw) + scale_x_continuous(breaks = seq(0,1200, 200))#, aes(y = ..density..),alpha = 0.5) #+
#   geom_histogram()+ scale_x_log10(breaks = seq(0,1200, 100))#(breaks = seq(0,1200, 200))
#     #geom_density(aes(y = ..density..), color = 'blue') +
#     #geom_rug()
#   print(p)
# }
# plot_hist(train)
# df = subset(train, loan_amount <=1200)
# df = df %>% 
#   mutate
# (p = ggplot(subset(train, state_code == "45"), aes(x = applicant_income,y = loan_amount)) + geom_point() + geom_smooth())
  
```

## **Data Recoding and Visualization**

Variable recoding and visualization are key steps that ensure every variable is of the desired class/type. Visualizaation is key to understand the distribution and separation of values by the two levels of a binary label.

### **Data Recoding**

Our data has a number of categorical featuers (`msa_md`, `state_code`,`county_code`,`lender`,`loan_type`,`property_type`, `loan_purpose`, `occupancy`,`preapproval`,`applicant_ethnicity`,`applicant_race`, `applicant_sex`) but which are read in as numeric. We begin by recoding such variables into the desired character class and explicitly supplying the categories in both train and test sets. We then  visualize the separation of values between different levels of the label - `accepted(0,1)`.

### **Data Visualization**

1. **Class separation by numeric variables**

We explore the distinction of the distribution of values between the two levels of the label through boxplots.

```{r Data viz-numerics, warning=FALSE, echo=FALSE, message=FALSE}
 num_vars<-sapply(names(train),function(x) ifelse(is.numeric(train[,x]) & x !="row_id",x,NA))
 num_vars<-na.omit(num_vars)

 cat_vars<-sapply(names(train),function(x) ifelse(is.character(train[,x]),x,NA))
 cat_vars<-na.omit(cat_vars)

# cat_vars = function(x) {
#   x = sapply(names(x), function(x) ifelse(is.factor(train[,x]),x,NA))
#   x = na.omit(x)
#   
# } 

#train[, numeric_vars(train)] = sapply(train[, numeric_vars(train)], scale)
#test_values[, numeric_vars(test_values)] = sapply(test_values[, numeric_vars(test_values)], scale)
#ss=apply(train[, sapply(train, is.numeric)],2, scale)
train$accepted = as.factor(train$accepted)
plot_box <- function(df, cols, col_x = 'accepted'){
  options(repr.plot.width = 4, repr.plot.height = 3.5)
  for (col in cols){
    p <- ggplot(df, aes_string(col_x, col)) + 
        geom_boxplot() +
        ggtitle(paste('Box plot of', col, '\n vs.', col_x))
    print(p)
     
  }
}

num_var = c("loan_amount", "applicant_income", "population", "minority_population_pct", "ffiecmedian_family_income", "tract_to_msa_md_income_pct", "number_of_owner.occupied_units", "number_of_1_to_4_family_units")
plot_box(train, num_var)
```

From the plots, three variables - loan_amount, minority_population_pct and tract_to_msa_md_income_pct show clear separation of values between the two levels. Next we explore this separation by categorical variables.

1. **Class separation by categorical variables**

We explore the distinction of the distribution of values between the two levels of the label through bar plots.


```{r Data viz, warning=FALSE, echo=FALSE, message=FALSE}
plot_bars = function(df, catcols){
  options(repr.plot.width = 6, repr.plot.height = 5)
  temp0 = df[df$accepted == 0,]
  temp1 = df[df$accepted == 1,]
  for(col in cat_cols){
    p1 = ggplot(temp0, aes_string(col)) + 
      geom_bar() + 
      ggtitle(paste('Bar plot of \n', col, '\n for rejected loan applications')) + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1))
    p2 = ggplot(temp1, aes_string(col)) + 
      geom_bar() + 
      ggtitle(paste('Bar plot of \n', col, '\n for approved loan applications')) + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1))
    grid.arrange(p1, p2, nrow = 1)
  }
}
cat_cols = c("loan_type","property_type","loan_purpose","occupancy", "preapproval", "applicant_ethnicity", "applicant_race", "applicant_sex", "co_applicant")
plot_bars(train, cat_cols)
rm(list = ls(all = T))
```

We observe the following:

1.	Some features such as property_type, loan_purpose, applicant_ethnicity, applicant_race and co_applicant have an almost significant different distribution of categories between the label categories.

2.	Others features such as preapproval, applicant_sex, occupancy show small differences, but these differences are unlikely to be significant.

Next we do feature engineering in readiness for modelling.


## **Feature Engineering**

Data preparation is an important step. We recoded loan_amount and applicant_income to categorical strings which help in distinguishing levels within the dataset and imputed missing numeric variables with means. 


Feature engineering and modelling were done in python

## **Model Building**

Modelling was done with python, selecting the CatBoostclassifier library for the classification. As earlier indicated we settled on 8 cross-validations with 1000 iterations to minimize bias and variance of the model. The classification model was defined as: 
`model_cat = CatBoostClassifier(iterations=1000, depth=8, learning_rate=0.4, eval_metric='Accuracy', use_best_model=True, random_seed=2019)`
and fitted as:
`model_cat.fit(xtrain, ytrain, cat_features = category_index, eval_set = (xtest, ytest))`
with best test of .72 and best iteration at 159



```{python, warning=FALSE, echo=FALSE, message=FALSE, results = 'hide'}
import pandas as pd
import numpy as np
from catboost import CatBoostClassifier as catclass
#from google.colab import files
import io



from sklearn.model_selection import train_test_split
from catboost import CatBoostClassifier, Pool, cv
#from google.colab import drive

#drive.mount("/content/gdrive", force_remount=True)

import os
#colab_path=[x[0]  for x in os.walk('/content/gdrive') if "Capstone Project" in x[0]];colab_path
colab_path=[x[0]  for x in os.walk('C:/Users/bokola/Analysis') if "capstone" in x[0]];colab_path
os.chdir(colab_path[3])

train_values = pd.read_csv("https://www.dropbox.com/s/3l8eyc51iupjglf/train_values.csv?dl=1")
test_values = pd.read_csv("https://www.dropbox.com/s/2heyrfbq3cxvsdb/test_values.csv?dl=1")
train_labels = pd.read_csv("https://www.dropbox.com/s/qvq3hg3mnlowxql/train_labels.csv?dl=1")

test_values.head(n=2)

train_labels.head(n=2)

train_values.head(n=2)

#Make rows Strings
train_values["row_id"]=train_values["row_id"].astype("object")
test_values["row_id"]=test_values["row_id"].astype("object")
train_labels["row_id"] = train_labels["row_id"].astype("object")

train_values["co_applicant"]=train_values["co_applicant"].astype("int64")
test_values["co_applicant"]=test_values["co_applicant"].astype("int64")

#train_values["minority_population"]=train_values["minority_population_pct"]*train_values["population"]/100
#test_values["minority_population"]=test_values["minority_population_pct"]*test_values["population"]/100

#train_values["tract_to_msa_md_income"]=train_values["tract_to_msa_md_income_pct"]*train_values["ffiecmedian_family_income"]/100
#test_values["tract_to_msa_md_income"]=test_values["tract_to_msa_md_income_pct"]*test_values["ffiecmedian_family_income"]/100

#Select Numeric Columns
numcols = ['population', 'minority_population_pct',
       'ffiecmedian_family_income', 'tract_to_msa_md_income_pct',
       'number_of_owner-occupied_units', 'number_of_1_to_4_family_units']

numcols1 = ['loan_amount','applicant_income']

for col in numcols1:
  bins_train  = [train_values[col].min(), train_values[col].quantile(.50),train_values[col].quantile(.75), train_values[col].max()]
  bins_test  = [test_values[col].min(), test_values[col].quantile(.50),test_values[col].quantile(.75), test_values[col].max()]

  train_values[col] = pd.cut(train_values[col],bins_train, labels =["1","2","3"]).astype("category")

  test_values[col] = pd.cut(test_values[col],bins_test, labels =["1","2","3"]).astype("category")

test_values.head()

train_values.head(n= 2)

for col in numcols1:
  train_values[col] = pd.to_numeric(train_values[col], errors='coerce')
  test_values[col] =  pd.to_numeric(test_values[col], errors='coerce')

#Replace Missing Values for numeric columns with mean of each
for numcol in numcols:
    train_values[numcol] = train_values[numcol].fillna(-999)

    test_values[numcol] = test_values[numcol].fillna(-999)



for numcol in numcols1:
    train_values[numcol] = train_values[numcol].fillna(-1)

    test_values[numcol] = test_values[numcol].fillna(-1)

    #train_values[numcol] = train_values[numcol].cat.add_categories("-1").fillna("-1")
    #test_values[numcol] = test_values[numcol].cat.add_categories("-1").fillna("-1")



train_values.isnull().sum()[0:10]

train_values.head(n=2)

test_values.head(n=2)

test_values_clean = test_values
train_values_clean =train_values

all_train = pd.merge(train_values_clean,train_labels, on="row_id", how="inner")

y_labs = train_labels["accepted"]


train_values_clean_1 = train_values_clean.loc[:, ~train_values_clean.columns.isin(['row_id'])]
test_values_clean_1 = test_values_clean.loc[:, ~test_values_clean.columns.isin(['row_id'])]

#Names of categorical columns in a list for looping on two datasets
catcols = ['loan_type', 'property_type', 'loan_purpose', 'occupancy','preapproval', 'msa_md', 'state_code', 'county_code',
       'applicant_ethnicity', 'applicant_race', 'lender','applicant_sex', 'co_applicant', 'loan_amount', 'applicant_income']

category_index =  [train_values_clean_1.columns.get_loc(c) for c in catcols if c!= "accepted"]; category_index

xtrain,xtest,ytrain,ytest = train_test_split(train_values_clean_1,y_labs,test_size=0.22)
model_cat = CatBoostClassifier(iterations= 1000, depth=8, learning_rate=0.4,eval_metric='Accuracy',use_best_model=True, random_seed=1920)

model_cat.fit(xtrain,ytrain,cat_features=category_index,eval_set=(xtest,ytest))

pred = model_cat.predict(test_values_clean_1)

import sklearn.metrics as metmod

def score_model(probs, threshold):
    return np.array([1 if x > threshold else 0 for x in probs[:,1]])

def key_metrics(labels, probs, threshold):
    scores = score_model(probs, threshold)
    metrics = metmod.precision_recall_fscore_support(labels, scores)
    conf = metmod.confusion_matrix(labels, scores)
    print('                 Confusion matrix')
    print('                 Score positive    Score negative')
    print('Actual positive    {}'.format(conf[0,0]) + '             {}'.format(conf[0,1]))
    print('Actual negative     {}'.format(conf[1,0]) + '             {}'.format(conf[1,1]))
    print('')
    print('Accuracy        {}'.format(round(metmod.accuracy_score(labels, scores), 2)))
    print('AUC              {}'.format(round(metmod.roc_auc_score(labels, probs[:,1]), 2)))
    print('Macro precision {}'.format(round(float((float(metrics[0][0]) + float(metrics[0][1]))/2.0), 2)))
    print('Macro recall    {}'.format(round(float((float(metrics[1][0]) + float(metrics[1][1]))/2.0), 2)))
    print(' ')
    print('           Positive      Negative')
    print('Num case   {}'.format(round(metrics[3][0], 2)) + '        {}'.format(round(metrics[3][1], 2)))
    print('Precision  {}'.format(round(metrics[0][0], 2)) + '         {}'.format(round(metrics[0][1], 2)))
    print('Recall      {}'.format(round(metrics[1][0],2)) + '        {}'.format(round(metrics[1][1], 2)))
    print('F1         {}'.format(round(metrics[2][0], 2)) + '         {}' .format(round(metrics[2][1],2)))

probabilities = model_cat.predict_proba(xtest)
key_metrics(ytest, probabilities, 0.5)

pred_df  = pd.DataFrame({'row_id':test_values.row_id ,'accepted': pred})
pred_df['accepted'] = pred_df.accepted.astype('int64')


pred_df.to_csv('predicted.csv', index=False)



```

# **Results and Conclusions**

The model was able to correctly classify 72% of the accepted values, but there were discripancies with the supplied test values indicating that our model was overfitting. However, it offers a good starting point to build on and improve.

                 Confusion matrix
                 Score positive    Score negative
Actual positive    36750             18402
Actual negative     12255             42593

Accuracy        0.72
AUC              0.8
Macro precision 0.72
Macro recall    0.72
 
           Positive      Negative
Num case   55152        54848
Precision  0.75         0.7
Recall      0.67        0.78
F1         0.71         0.74















