---
title: "data.table"
author: "Basil Okola: [`@okolaowiti`](https://twitter.com/okolaowiti)"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    nature: 
      beforeInit: "https://platform.twitter.com/widgets.js"
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
      ratio: '16:9'
      self_contained: true
    css: xaringan-themer.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE)
```

```{r libs, message=FALSE, results='hide'}
ipk = function(pkg){
  new.pkg = list.of.pkgs[!(list.of.pkgs %in% .packages(all.available = TRUE))]
  if(length(new.pkg)) install.packages(new.pkg, dependencies = TRUE)
  suppressPackageStartupMessages(sapply(pkg, library, character.only = TRUE))
}

list.of.pkgs = c('data.table', 'tidyverse')
ipk(list.of.pkgs)

```

```{r xaringan-themer, eval = FALSE, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light(
  base_color = "#374785",
  text_color = darken_color("#8E8D8A", 0.5),
  white_color = lighten_color("#C2CAD0", 0.95),
  code_inline_color = choose_dark_or_light("#E98074", "#E85A4F"),
  header_font_google = google_font("Noto Sans"),
  text_font_google   = google_font("Karla", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  extra_css = list(".hljs-github .hjls" = list("background" = "black"),
                   ".small" = list("font-size" = "70%"),
                   ".small pre .remark-code" = list("font-size" = "75%"),
                   ".small .remark-code" = list("font-size" = "75%"),
                   ".tiny .remark-code" = list("font-size" = "50%"),
                   ".medium .remark-code" = list("font-size" = "90%")
  )
)
```

```{css, include = FALSE, eval = TRUE}
/* Extra CSS */
.hljs-github .hjls {
  background: black;
}
.small {
  font-size: 70%;
}
.small pre .remark-code {
  font-size: 75%;
}
.small .remark-code {
  font-size: 75%;
}

.tiny {
  font-size: 60%;
}
.tiny pre .remark-code {
  font-size: 60%;
}
.tiny .remark-code {
  font-size: 60%;
}

.tiny .remark-code {
  font-size: 60%;
}

.medium {
  font-size: 80%;
}
.medium pre .remark-code {
  font-size: 85%;
}
.medium .remark-code {
  font-size: 85%;
}

.remark-slide thead, .remark-slide tr:nth-child(2n) {
  background-color: white;
}

.remark-slide thead, .remark-slide tr:nth-child(2n) {
  background-color: white;
}

/* footer */

div.my-footer {
    background-color: #1a1917;
    position: absolute;
    bottom: 0px;
    left: 0px;
    height: 20px;
    width: 100%;
}

div.my-footer span {
    font-size: 10pt;
    color: #F7F8FA;
    position: absolute;
    left: 15px;
    bottom: 2px;
}

```

---
You can use `expand.grid` to create a dataframe <br>

```{r}
expand.grid(pos = c("interior", "middle", "exterior"),
dose = c(100, 150),
type = c("WT","KO"))
```


---
Data table syntax is of the form `DT[i, j, by]`

- `i`: on which row
- `j`: what to do
- `by`: group by what


# Data Manipulation

- Just looking
- Removing information
- Adding information
- Reducing information
- Combining information

---
## Just looking

```{r}
irisDT = as.data.table(iris)
tables() # show loaded tables
```



Sorting / ordering rows

setorder(data.table, ...)

`-`: to sort a variable in descending order

---


```{r}
setorder(irisDT, Sepal.Length, Sepal.Width)
irisDT
```


```{r}
setorder(irisDT, -Species, Sepal.Width)
irisDT
```


---

## Removing information

### Selecting rows

DT[i, j, by]

- `i`: on which rows?

```{r}
irisDT[3:4]
```

```{r}
irisDT[3:4,] #row 3 and 4
irisDT[-(1:5)] #delete rows 1:5
irisDT[!(1:5)] # using false to delete as well
irisDT[.N] # last row
irisDT[1:(.N-10)] # all but the last 10 rows
irisDT[Species == "virginica"] # based on conditions fulfilled in columns
irisDT[Species %like% "^v"] # using %like% helper function
irisDT[Petal.Width %between% c(0.3, 0.4)] # values in an interval
```
---

### Selecting columns

DT[i, j, by]

- `j`: what to do? -> select columns

```{r}
irisDT[, Species] # returns a vector
irisDT[, "Species"] # returns a dataframe/data.table
irisDT[, -c("Species")]
irisDT[, !c("Species")]
irisDT[, list(Species, LS = Sepal.Length)] # select and rename
# .() is an alias for list

irisDT[, .(Species, SL = Sepal.Length)]
irisDT[Species == "virginica" & Sepal.Length > 7 & Sepal.Width < 3, !c("Species")]
irisDT[Species == "virginica" &
Sepal.Length > 7 &Sepal.Width < 3,.(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width)]
```
---
## Adding information

### Making new columns - preserving existing ones

DT[i, j, by]

- `j`: what to do? -> compute new columns preserving existing ones

data.table uses a new operator `:=` to add/update/delete columns (by reference)
---
1. option 1

```{r}
irisDT[, maxLength := max(Sepal.Length, Petal.Length)][,
          minWidth := min(Sepal.Length, Petal.Width)]
irisDT
```
---
2. option 2

LHS := RHS form

```{r}
irisDT[, c("maxLength", "minWidth") :=
         list(
           max(Sepal.Length, Petal.Length),
           min(Sepal.Width, Petal.Width)
         )]

irisDT
```
---
3. Option 3

Functional form

```{r}
irisDT[, `:=` (
  maxLength = max(Sepal.Length, Petal.Length),
  minWidth = min(Sepal.Width, Petal.Width)
)]
irisDT
```
---
### Making new columns - dropping existing ones

DT[i, j, by]

- `j`: what to do? -> compute new columns dropping existing ones

```{r}
irisDT[, .(maxLength = pmax(Sepal.Length, Petal.Length),
minWidth = pmin(Sepal.Width, Petal.Width))]
```

```{r}
newIris <- irisDT[, .(Sepal.Area = Sepal.Width * Sepal.Length,
Petal.Area = Petal.Width * Petal.Length)]
newIris[, Area.Ratio := Petal.Area / Sepal.Area]
newIris
```
---

## Reducing information

### Summarizing rows

```{r}
irisDT[, .(meanSepalLength = mean(Sepal.Length),
           meanSepalWidth = mean(Sepal.Width),
           meanPetalLength = mean(Petal.Length),
           meanPetalWidth = mean(Petal.Width))]
```
```{r}
irisDT[, .(nSamples = .N, nSpecies = uniqueN(Species))
       ]
```
---

### Summarizing rows with filtering

```{r}
irisDT[Species == "versicolor",.(
  meanSepalLength = mean(Sepal.Length),
           meanSepalWidth = mean(Sepal.Width),
           meanPetalLength = mean(Petal.Length),
           meanPetalWidth = mean(Petal.Width)
)]

```

---

### Grouping by one or more variables

```{r}
irisDT[, .(meanSepalLength = mean(Sepal.Length),
meanSepalWidth = mean(Sepal.Width),
meanPetalLength = mean(Petal.Length),
meanPetalWidth = mean(Petal.Width)), by = "Species"]
```

```{r}
irisDT[, .(nSamples = .N, nSpecies = uniqueN(Species)), by = .(Species)]
```

---

## Combining information

### joining 2 data.tables

**Inner join**: return all rows from x where there are matching values in y and
all columns from x and y. In case of multiple matches between x and y, all
combinations of the matches are returned.<br>

**Full join**: return all rows and columns from both x and y. <br>

**Left join**: return all rows from x and all columns from x and y. In case of
multiple matches between x and y, all combinations of the matches are
returned.<br>

**Right join**: return all rows from y and all columns from x and y. In case of
multiple matches between x and y, all combinations of the matches are
returned.<br>
---

DT[i, on]

- `i`: join to which data.table?
- `on`: join key columns?

```{r}
(x <- data.table(id = c(1, 2, 4, 5, 6),
x = c(9, 12, 14, 21, 8)))

(y <- data.table(id = c(1, 3, 4, 6, 6),
y = c(8, 14, 19, 2, 4)))
```
### inner join - data.table

```{r}
y[x, on = .(id), nomatch = 0]
```
---

### full join - merge

```{r}
merge.data.table(x = x, y = y, by = "id", all = TRUE)
```


### left join - merge

```{r}
merge.data.table(x = x, y = y, by = "id", all.x = TRUE)
```
#### left join - data.table

```{r}
y[x, on = .(id)]
```
---

### right join - merge

```{r}
data.table::merge.data.table(x = x, y = y, by = "id", all.y = T)
```

### right jpin - data.table

```{r}
x[y, on = .(id)]
```

### antijoin - data.table

```{r}
x[!y, on = .(id)]
```
---

# Keys

No need of the on argument when performing a join

- Sorts the data.table in memory by the key column(s)
- Multiple columns can be set and used as keys
- Useful functions:
- setkey(DT, ...) or setkeyv(DT, keys)
- haskey(DT)
- key(DT

```{r}
setkey(x, id)
setkey(y, "id")
haskey(x); haskey(y)
key(x); key(y)
```

---
## joins with keys already specified

```{r}
# Inner join x and y
x[y, nomatch = 0]
# Right join x and y
x[y]
# Left join x and y
y[x]
```
---

## Exercise 3

```{r}
library(dplyr)
library(data.table)
data(starwars)
starwarsDT <- as.data.table(starwars)
starwarsDT[1]
starwarsDT[, .(min = min(mass, na.rm = T),
               max = max(mass, na.rm = T),
               nobs = .N), by = homeworld][nobs >1] 
```
---
# Reshaping data with data.table


---


- Tidy data

  - Each variable is a column
  -Each observation is a row
  - Each value is a cell

- Two fundamental verbs
  - `melt()`: 'Wide' -> 'long' data
  - `dcast()`: 'long' -> 'wide' data
  
---

**Wide format**

```{r}
knitr::kable(head(iris[1:4,], n =3))
```

`melt(wide_df, id.vars = 'grouping variable', variable.name = 'measures_across_grouping_var',value.name = 'value_name_as_desired')`
---
 
 
**Long to wide**

`dcast(long_df, group ~ atttribute, value.var = "value")`

---

# Separating and uniting information

```{r}
irisUnited <- irisDT[ , Sepal.Length.Width := paste0(Sepal.Length, "/", Sepal.Width)]
irisUnited[1:5]
```

```{r}
irisSeparated = irisUnited[, c('Sepal.Length', 'Sepal.Width') := tstrsplit(Sepal.Length.Width, "/")]
irisSeparated[1:5]
```


---

# Hands-on session


Exercise 1

```{r}
data("relig_income")
a = as.data.table(relig_income)
a[1:10]
b = melt(a, id.vars = "religion", variable.name = "income_category", value.name = "n")
```


Exercise 2

```{r}
data("storms")
aa = as.data.table(storms)
aa[, date:= paste0(year, "-", month,"-", day)]
```


---

# Useful symbols/ functions

- `.SD`: Subset of Data for each group
  - .SD along with lapply can be used to apply any function to multiple columns by group

- `.SDcols`: Columns of the data.table that are included in .SD
- na.omit(): returns the object with incomplete cases removed

---


## Summarizing for specific condition

```{r}
cols = sapply(irisDT, is.numeric)
cols = names(cols)[cols]
irisDT[, lapply(.SD, mean), by = Species, .SDcols = cols]
```

---

## Summarizing for specific variable of interest

```{r}
irisDT[, lapply(.SD, mean), by = Species, .SDcols = c("Petal.Length", "Petal.Width")]
```

---

## Summarizing all columns

```{r}
irisDT[ , unlist(recursive = FALSE, lapply(.(mean = mean, sd = sd, min = min, max = max), 
                                           function(f) lapply(.SD, f))), by = Species]
```

---

# Hands-on 3

Exercise 1 - Compute the minimum, maximum, median and mean for all numeric columns

```{r}
data("starwars")
starwarsDT = as.data.table(starwars)
cols = sapply(starwarsDT, is.numeric)
cols = names(cols)[cols]

starwarsDT[, height:= as.numeric(height)]

starwarsDT[, unlist(recursive = F,
                    lapply(.(min = min, max = max, median = median, mean = mean), function(f) lapply(.SD, f))), by = .(homeworld, species), .SDcols = cols]
```

---

Exercise 2 - all colnames to upper case

```{r}
starwarsDT = as.data.table(starwars)
setnames(starwarsDT, toupper(names(starwarsDT)))
```

Convert all character columns to upper case

```{r}
cols = sapply(starwarsDT, is.character)
cols = names(cols)[cols]
starwarsDT[, lapply(.SD, toupper)]
```

---

Exercise 3

- Remove the data which have NA for the numeric variables
- Group the data by all character variables
- Compute the mean for `ts_diameter` and `hu_diameter`

```{r}
stormsDT <- as.data.table(storms)
numericCols = sapply(stormsDT, is.numeric)
numericCols = names(numericCols)[numericCols]
stormsDTNew <- na.omit(stormsDT, cols = numericCols)
characterCols <- sapply(stormsDTNew, is.character)
characterCols <- names(characterCols)[characterCols]
colsOfInterest <- c("ts_diameter", "hu_diameter")
stormsDTNew[, lapply(.SD, mean), .SDcols = colsOfInterest, by = characterCols]
```

---

`.SD`: Subset of data - choose row by ordinal position

```{r}
irisDT[, .SD[1L]]
irisDT[, .SD[1L], by = Species]
irisDT[, .SD[1L], by = Species, .SDcols = c("Petal.Length", "Petal.Width")]
```

---

`.SD`: Subset of Data - select top entries (in each group)

```{r}
irisDT[order(-Sepal.Length), .SD[1:2]]
irisDT[, .SD[which.max(Sepal.Length)], by = Species, .SDcols = c("Sepal.Length","Sepal.Width")]
```

---

# Multiple if else statements

```{r}
irisDT[, text := {
  if(Sepal.Length > 4 & Sepal.Width > 4) "sepal length and width is larger than 4"
  else if(Petal.Length < 2 & Petal.Width < 2) "petal length and width is smaller than 2"
  else "other"
}, by = .(Species, Sepal.Length, Sepal.Width, Petal.Length, Petal.Width)][, .SD[1], by = text]
```


```{r}
bin_iris <- function(sl, sw, pl, pw){
  if(sl > 4 & sw > 4) "sepal length and width is larger than 4"
  else if(pl < 2 & pw < 2) "petal length and width is smaller than 2"
  else "other"
}
irisDT[, text := bin_iris(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width), 
       by = .(Species, Sepal.Length, Sepal.Width, Petal.Length, Petal.Width)][, .SD[1], by = text]
```

---

# Hands - on 4

Exercise 1


Print the fifth row for each species.

```{r}
starwarsDT <- as.data.table(starwars)
starwarsDT[, if(.N >= 5) .SD[5L], by = species]
```

---


Exercise 2

- Consider the columns name, status and pressure.
- Print for each storm, the 5 entries with the lowest pressure.

```{r}
stormsDT <- as.data.table(storms)
stormsDT[, .(name, status, pressure)][order(name, pressure)][, .SD[1:min(5, .N)], by = name]
```


