---
title: "External data in R"
author: "  \nBasil Okola"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2: default
  pdf_document: default
  bookdown::word_document2: default
  html_document:
    df_print: paged
tables: yes
fontsize: 14pt
mainfont: Arial
fig_caption: yes
number_sections: yes
toc: yes
header-includes: \usepackage{amsmath}
#bibliography: citation.bib
---

```{r, child="_setup.Rmd"}

```


```{r opts, include=FALSE}
home = ifelse(Sys.info()[["sysname"]] == "Linux", Sys.getenv("HOME"), Sys.getenv("USERPROFILE"))
home = gsub("\\\\", "/",home)

data_dir = file.path(home, "Google Drive (basil.okola@student.uhasselt.be)", "MSc. Stats Hasselt", 
                     "y1 sem1", "Linear models")
fig_dir = file.path(home, "Google Drive (basil.okola@student.uhasselt.be)", "MSc. Stats Hasselt", 
                    "y1 sem1", "Linear models")
```


```{r packages, include=FALSE, echo=FALSE}
ipk = function(pkg){
  new.pkg = list.of.pkgs[!(list.of.pkgs %in% .packages(all.available = TRUE))]
  if(length(new.pkg)) install.packages(new.pkg, dependencies = TRUE)
  suppressPackageStartupMessages({sapply(pkg, require, character.only = TRUE)})
}

list.of.pkgs = c("tidyverse", "magrittr", "rvest", "httr", "jsonlite", "DBI", "config", "vaultr", "dbplyr")
ipk(list.of.pkgs)

```

\newpage

# Web scraping

## rvest package

Harvest/scrape data from web:

- `read_html()`: collects data from the webpage
- `html_nodes()`: extract the relevant pieces
- `html_text()`: extract tags of the relevant piece
- `html_attributes()`: extract attributes of the relevant piece

**Read HTML code**

```{r, warning=FALSE, message=FALSE}
url = 'https://www.imdb.com/title/tt6723592/'
# reading the html code from website
webpage = read_html(url)
# using the html tag
castWithTag = html_nodes(webpage, "div")
length(castWithTag)
# using the id attribute, note the '#'
castWithId_node = html_nodes(webpage, "#titleCast")
html_text(castWithId_node)
# using the class attribute: note the '.'
castWithClass_node = html_nodes(webpage, ".cast_list")
html_text(castWithClass_node)
# using a more detailed attribute
cast = html_nodes(webpage, ".primary_photo + td a")
html_text(cast)
```

**Extracting a table**

```{r, message=FALSE, warning=FALSE}
allTables = html_table(webpage, "table", header = FALSE)
castTable = html_table(webpage, ".cast_list", header = FALSE)[[1]]
head(castTable)
```


**Attributes of an element**
```{r, message=FALSE, warning=FALSE}
html_attrs(castWithClass_node)
html_attr(cast, "href")
```

# API Requests

R packages: `httr` and `jsonlite`

## httr

- create the request with GET() function
- input is a url, which specifies the address of the server

- Example: current number of people in space

```{r, warning=FALSE, message=FALSE}
res = GET("http://api.open-notify.org/astros.json")
res
```

**JSON Format**
JSON = JavaScript Object Notation

## jsonlite

- The `jsonlite` package provides parser and generator functions: `fromJSON()`, `toJSON()`

- `fromJSON()`:
  - Input is a JSON string, URL, JSON file
  - Retruns a list of data.frames

- `toJSON()`:
  - input is any object
  - returns a JSON string
  
### fromJSON

- convert the raw unicode of the GET request into a JSON string

```{r, warning=F, message=F}
rawToChar(res$content)
```

- convert the JSON format to a list of data.frames

```{r, message=F, warning=F}
data = fromJSON(rawToChar(res$content))
data
```

- Read directly from a url e.g github page <br>

```{r, warning=F, message=F}
data <- fromJSON("https://api.github.com/users/hadley/orgs")
class(data)
```

### toJSON

- write a data.frame to a JSON format

```{r, warning=F, message=F}
jsonData = toJSON(data)
head(jsonData)
```

- and back from the JSON format to data.frame

```{r, warning=F, message=F}
backAgain = fromJSON(jsonData)
class(backAgain)
identical(data, backAgain)
```
## API with query parameters

- query param: longitude, latitude

```{r, message=F, warning=F}
res = GET("http://api.open-notify.org/iss-pass.json",
query = list(lat = 40.7, lon = -74))
res

```

# Databases

## Connecting

- Find a [DBI compliant package] (https://cran.r-project.org/web/views/Databases.html) that implements the driver <br>
- After you are finished using the connection, you should always disconnect.<br>

```{r, warning=F, message=F}
conn = dbConnect(
  dbname = file.path(home, 'Downloads', 'chinook.db'),
  drv = RSQLite::SQLite()
)

# DBI::dbDisconnect(conn)
```

### connecting with credentials

```{r, eval=F, message=F, warning=F}
connn <- DBI::dbConnect(
user = "oa",
dbname = "oaDB",
password = "oaPassword",
host = "oaHost",
port = "1111",
drv = RPostgres::Postgres())
DBI::dbDisconnect(connn)
```

- Credentials are sensitive data
- better not hard-code credentials
- A solution is the use of environment variables

- interactively set environment variables

```{r, warning=F, message=F}
Sys.setenv("DB_USER" = "bokola")
Sys.getenv('DB_USER')
```

- Save environment variables permanently by setting them in the .Renviron file:
  - Loaded at the beginning of a session
  - Overruled by setting the environment variable with `Sys.setenv`

`usethis::edit_r_environ()`

DB_USER = "myNewUser"

- Or use your IDE (Rstudio, Architect,…) to specify them on startup of a session

### Connect with Environment Variables


```{r, eval=FALSE}
connectDB = function(
  user = Sys.getenv('DB_USER'),
  dbname = file.path(home, 'Downloads', 'chinook.db'),
  password = Sys.getenv("DB_PASSWORD"),
  host = Sys.getenv("DB_HOST", "localhost"),
  port = Sys.getenv("DB_PORT", "5432"),
  drv = RSQLite::SQLite()
){
  args = as.list(environment())
  do.call(dbConnect, args)
}
conn = connectDB()
```

### Credential managing packages

`config`, `vaultr`

## Retrieving data

DBI: dbReadTable(), dbListFields(), dbListTables()


```{r, warning=FALSE, message=FALSE}
dbListTables(conn)
dbListFields(conn, "Artists")
dbReadTable(conn, "Artists") %>% head(3)
```

## dbplyr

- For basic queries, a good option is the `dbplyr` package
- No SQL knowledge required and learn some SQL while using the package

```{r, warning=FALSE, message=FALSE}
Artists = tbl(conn, "Artists")
Artists %>%
  filter(Name %in% "AC/DC") %>%
  show_query()
```

## DBI: dbGetQuery

- Assemble your SQL queries and pass these to dbGetQuery()
- Best use: only for SHOW and SELECT statements

```{r, warning=FALSE, message=FALSE}
dbGetQuery(conn, "select * from artists limit 3")

```

## Compaarison: Where to do the data wrangling

Example: selecting 3 most expensive tracks

```{r}
# base r

df = dbReadTable(conn, "tracks")
df = df[order(df$UnitPrice, decreasing = TRUE),]
df[1:3,]
```

```{r}
# data.table
df = dbReadTable(conn, "tracks")
DT = data.table::data.table(df)
DT[order(-UnitPrice)][1:3]
```

```{r}
# dplyr
df = dbReadTable(conn, "tracks") %>%
  arrange(desc(UnitPrice)) %>%
  head(n = 3)
df
```

```{r}
# SQL
df = dbGetQuery(conn, "select * from tracks order by UnitPrice desc limit 3")
df
```

## DBI functions

- provide data to database using `dbExecute()` and/or `dbAppendTable()`
- Prefer not to use dbWriteTable(), since this can overwrite the schema. Use instead a proper database
management tool in setting up new tables.

```{r}
dbAppendTable(
  conn = conn,
  name = "Artists",
  value = data.frame(ArtistId = "876", Name = "newBand")
)
```

```{r}
dbExecute(conn,
          "insert into Artists (Name) VALUES ('newBandnewName');")
```

```{r}
dbExecute(conn,
          "delete from Artists where name = 'newBand';")
```

## More DBI functions

- Other DBI functions that can be used to store/update/delete data
- `dbExecute()` should be preferred

```{r, eval=F}
dbAppendTable(
  conn = conn,
  name = "Artists",
  value = data.frame(
ArtistId = c("867", "868", "869"),
Name = c("newBand1", "newBand2","newBand3"))
)
```


Using parametarized query multiple times

```{r}
res = dbSendStatement(conn, "delete from Artists where Name = $band;")
dbBind(res, list(band = 'newBand'))
dbGetRowsAffected(res)
dbClearResult(res)
```

## How do I assemble queries

Avoid using sprintf() and paste() as much as possible!

```{r}
res = DBI::dbExecute( # Same for dbSendQuery, dbGetQuery
  conn = conn,
  statement = "delete from Artists where Name = $1;",
  params = 'newBand'
)

myQuery = DBI::sqlInterpolate(
  conn = conn,
  sql = "delete from Artists where Name = ?name;",
  name = "newBand"
)

myQuery = glue::glue_sql(
  .con = conn,
  "delete from Artists where Name in ({names*})",
  names = c("newBand", "newBand2") # 1 placeholder, variable length argument!
)

print(myQuery)
res = dbExecute(conn, myQuery)
```

## Single connection vs pooling


[pool](https://cran.r-project.org/web/packages/pool/index.html) is a package that manages your connections


```{r, eval=FALSE}
pool <- pool::dbPool(
user = "oa",
dbname = "oaDB",
password = "oaPassword",
host = "oaHost",
port = "1111",
drv = RPostgres::Postgres())
conn <- poolCheckout(pool)
## some code using the connection
poolReturn(conn)
pool::poolClose(pool)
```







