---
title: "Neonatal admission trends from routinely collected observational data"
author: "bokola"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::word_document2:
    fig_caption: yes
    toc: yes
    toc_depth: '3'
  bookdown::pdf_document2:
    latex_engine: lualatex
    toc: yes
    toc_depth: '3'
  bookdown::html_document2:
    highlight: tango
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 3
---


```{r setup , include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      fig.align = "center",
                      fig.width = 6.5, 
                      fig.height = 4, 
                      dev = "png",
                      cache = FALSE)
options(tinytex.verbose = TRUE)
```

This is an analysis focusing on understanding if there has been any changes in admission numbers in the neonatal wards across participating CIN sites. 

# METHODS

This study utilizes seanonal-trend decomposition to analyse trend in admission of neonates in select public hospitals in Kenya.

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide'}
# Data retrieval and preparations
rm(list = ls(all=T))


rpkgs <- c("rlang", "tidyselect", "gWidgets","RGtk2","gWidgetsRGtk2","data.table","RCurl","zoo","RedcapData",
           "devtools", "ReporteRs","stringr", "dplyr","grid","ggplot2","zoo", "lubridate", "digest", "janitor", "gt", "ts")

if(any(!rpkgs %in% .packages(all.available = TRUE))){
  rpkgs1<-rpkgs[(!rpkgs %in% .packages(all.available = TRUE))]
  rpkgs1 <- rpkgs1[!rpkgs1 %in% "RedcapData"]
  install.packages(rpkgs1,dep=TRUE,repos='http://cran.rstudio.com')
}
#Redcap development packages
# if(!'RedcapData' %in% .packages(all.available = TRUE)){
#   devtools::install_github('tjhubs/RedcapData',dep=TRUE,ref='v7fixed')
# }
lapply(rpkgs[(rpkgs %in% .packages(all.available = TRUE))], library, character.only = TRUE)

#source(file.path(home.dir,"hsuApps", "universalFunctions", "modifyData.R"))

if(tolower(Sys.info()[['sysname']])=="linux"){
  home.dir <- normalizePath(paste0(Sys.getenv()[["HOME"]],"/hsuApps")) %>% sub("\\\\", "/", .)
}else{
  home.dir <- normalizePath(paste0(Sys.getenv()[["USERPROFILE"]],"/Analysis/hsuApps")) %>% gsub("\\\\", "/", .)
}

proj_dir = file.path(home.dir, 'neonatal analysis project')
data_dir = file.path(proj_dir, 'data')
cache_dir = file.path(proj_dir, "cache", "doc")
src_dir = file.path(proj_dir, "src", "r")

for(i in c(proj_dir, data_dir, cache_dir, src_dir)){
  if(!dir.exists(i)){
  if(dir.create(i, recursive = T))
    stop("Project directory\"", i, 
         "\"has been created!\nPlease fill it with relevant files")
  else
    stop("The directory", i, "could not be created")
  
  }
}


```
Loading and cleaning data

```{r,message=FALSE, warning=FALSE, echo=FALSE, results='hide'}

data = readr::read_csv(file.path(data_dir, "los_neodata2020-03-23.csv"))

dataXXX = data
source(file.path(src_dir, "time.R"))

t_cols <- grep("\\bt_adm\\b|\\bt_discharge\\b", names(data), v = T)
#cols = cols[-c(1,6,12,13,14,19,21, 27,29)]
for (c in t_cols) {
  data[, c] <- sapply(data[, c], ValidTime) #time columns (a[,c])
  
}

cols <- grep("diag", names(data), v = T)
#cols = cols[-c(1,6,12,13,14,19,21, 27,29)]
for (c in cols) {
  data[, c] <- sapply(data[, c], diag_tab) #diag_tab(a[,c])
  
}

data = data %>%
  mutate(
    .,
    birth_wt = ifelse(
      as.numeric(birth_wt) > 100 &
        !is.na(as.numeric(birth_wt)),
      as.numeric(birth_wt) / 1000,
      birth_wt
    ),
    hosp_id = gsub('/', '-', hosp_id),
    discharge_weight = ifelse(
      !is.na(discharge_weight) &
        as.numeric(discharge_weight) > 100,
      as.numeric(discharge_weight) / 1000,
      discharge_weight
    ),
    discharge_weight,
    wt_now = ifelse(
      !is.na(wt_now) &
        as.numeric(wt_now) > 100,
      as.numeric(wt_now) / 1000,
      wt_now
    ),
    wt_now = ifelse((is.na(wt_now) |
                       grepl("-1", data$wt_now)) &
                      date_adm == date_of_birth, birth_wt, wt_now)
  )

data = data %>%
  mutate(
    .,
    wt_now = ifelse((is.na(wt_now) |
                       grepl("-1", data$wt_now)) &
                      date_adm == date_of_birth, birth_wt, wt_now),
    birth.weight.group = ifelse(
      is.na(birth_wt) | birth_wt < 0,
      'Unrecorded',
      ifelse(
        birth_wt <= 1,
        '1000 gms & Below',
        ifelse(
          birth_wt < 1.5,
          '1001 - 1499 gms',
          ifelse(
            birth_wt < 2.5,
            '1500 - 2499 gms',
            ifelse(birth_wt < 4, '2500 - 3999 gms', "4000 gms & Above")
          )
        )
      )
    ),
    temp.group = ifelse(
      is.na(temperature_degrees_celciu) |
        as.numeric(temperature_degrees_celciu) < 0,
      'No Documentations',
      ifelse(
        temperature_degrees_celciu < 32,
        'l_32',
        ifelse(
          temperature_degrees_celciu <= 35.9 &
            temperature_degrees_celciu >= 32,
          'l_36',
          ifelse(
            temperature_degrees_celciu <= 36.4 &
              temperature_degrees_celciu >= 36.0,
            'l_36.5',
            ifelse(
              temperature_degrees_celciu <= 37.5 &
                temperature_degrees_celciu >= 36.5,
              'l_37.5',
              "m_37.5"
            )
          )
        )
      )
    )
  )
data = data %>%
  mutate(.,
         los = ifelse(
           !is.na(t_adm) &
             !is.na(t_discharge) &
             !is.na(date_adm) &
             !is.na(date_discharge) &
             !grepl('1913|1914|1915', date_adm) &
             !grepl('1913|1914|1915', date_discharge),
           round(as.numeric(
             difftime(
               strptime(paste0(date_discharge, t_discharge), "%Y-%m-%d %H:%M"),
               strptime(paste0(date_adm, t_adm), "%Y-%m-%d %H:%M"),
               units = "days"
             )
           )),
           ifelse(
             is.na(t_adm) |
               is.na(t_discharge) &
               !grepl('1913|1914|1915', date_adm) &
               !grepl('1913|1914|1915', date_discharge),
             round(as.numeric(
               difftime(
                 strptime(date_discharge, "%Y-%m-%d"),
                 strptime(date_adm, "%Y-%m-%d"),
                 units = "days"
               )
             )),
             NA
           )
         )) %>%
  mutate(
    .,
    los = ifelse(los < 0, NA, los),
    all.discharge.diag := paste(
      primary_disch_diagnosis,
      disch_diag_1,
      disch_diag_2,
      disch_diag_3,
      disch_diag_4,
      disch_diag_5,
      other_discharge_diag_1,
      other_discharge_diag_2,
      other_discharge_diag_3,
      other_discharge_diag_4,
      other_discharge_diag_5,
      other_disch_diag_old,
      other_discharge_diag_unlisted,
      sep = ';'
    )
  ) %>%
  mutate(., all.discharge.diag = diag_tab(all.discharge.diag)) %>%
  mutate(., all.discharge.diag = stringr::str_trim(all.discharge.diag)) %>%
  mutate(
    .,
    all.adm.diag = paste(
      pry_adm_diag,
      adm_diag_1,
      adm_diag_2,
      adm_diag_3,
      other_admission_diag_1,
      other_admission_diag_2,
      other_admission_diag_3,
      other_admission_diag_4,
      other_admission_diag_5,
      admisn_diag_not_listed,
      sep = ';'
    )
  ) %>%
  mutate(., all.adm.diag = diag_tab(all.adm.diag)) %>%
  mutate(., all.adm.diag = stringr::str_trim(all.adm.diag)) %>%
  mutate(., all.diag = paste(all.discharge.diag, all.adm.diag, sep = ';')) %>%
  filter(!is.na(date_discharge) & !grepl('1913|1914|1915', date_discharge)) %>%
  arrange(., date_discharge)

  wrong = data %>%
    filter(., date_discharge > Sys.Date()) %>%
    select(., c(id, hosp_id, date_discharge))

adm.diag = splitstackshape::cSplit(data, 'all.adm.diag', ';', direction = 'long') %>%
  mutate(., all.diag = as.character(all.diag)) %>%
  distinct(id, all.diag, .keep_all = TRUE) %>%
  subset(.,!all.diag == '')

disch.diag = splitstackshape::cSplit(data, 'all.discharge.diag', ';', direction = 'long') %>%
  mutate(., all.diag = as.character(all.diag)) %>%
  distinct(id, all.diag, .keep_all = TRUE) %>%
  subset(.,!all.diag == '')

all.diag = splitstackshape::cSplit(data, 'all.diag', ';', direction = 'long') %>%
  mutate(., all.diag = as.character(all.diag)) %>%
  distinct(id, all.diag, .keep_all = TRUE) %>%
  subset(.,!all.diag == '')

monthly_disch = data %>%
  filter(., date_discharge <= Sys.Date() &
           !grepl('2012', date_discharge)) %>%
  mutate(., month = as.Date(paste0(format(
    date_discharge, "%Y-%m"
  ), '-01'))) %>%
  group_by(month) %>%
  summarise(., discharges = n())

# ts object

discharges_ts = ts(monthly_disch$discharges, start = c(2013,1), freq = 12)


```

```{r, ts-plot, fig.cap='(ref:ts-plot)'}
knitr::opts_chunk$set(chunk_option1 = TRUE)

plot.ts(discharges_ts)
```

Figure \@ref(fig:ts-plot) indicates an increasing size of the seasonal fluctuations and random fluctuations rulling out utility of additive models in describing the number of discharges time series. We may then try and do a log transform of the time series and see if it can be described by an additive model <br>

```{r, log-ts-plot, fig.cap='(ref:log-ts-plot)'}
knitr::opts_chunk$set(chunk_option1 = TRUE)

log_discharges_ts = log(discharges_ts)
plot.ts(log_discharges_ts)

```






