---
output:
  bookdown::pdf_document2: default
  pdf_document: default
tables: yes
---

## Referencing tables and figures

- set output to bookdown options
- for tables do not set chunk name - set it in caption with `\\label{tab:chunk_name}`
- these labels must be unique
- use `format = pandoc`
- With `format='pandoc'` you need to enter the \label command in the caption.
- ref tables with \ref{tab:label_name} and
- figures with \@ref(fig:chunk_name)


```{r , results='markup'}
tab <- head(iris)
knitr::kable(tab, 
  format='pandoc', 
  digits = 3, 
  caption = "estimates and respective confidence intervals for a model incluiding the fastfood dummy variable\\label{tab:lat_table}"
  )
```

```{r, results='markup'}
tab <- head(iris)
knitr::kable(tab, 
  format='pandoc', 
  digits = 3, 
  caption = "estimates and respective confidence intervals for a model incluiding the fastfood dummy variable\\label{tab:pandoc_table}"
  )
```

```{r plot, fig.cap= "plot", results="markup"}
plot(iris$Sepal.Length)
```

Table \ref{tab:pandoc_table} was done using Pandoc,
and Figure \@ref(fig:plot) is done with base R

