---
title: "RunniRng SAS from Rstudio"
author: "Basil Okola"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
  bookdown::pdf_document2: default
  pdf_document: default
  bookdown::word_document2: default
  fig_caption: yes
  word_document: default
tables: yes
fontsize: 14pt
mainfont: Arial
fig_caption: yes
number_sections: yes
toc: true
header-includes: \usepackage{amsmath}
#bibliography: citation.bib SAS from Rstudio
---

```{r setup, child="_setup.Rmd"}
# knitr::opts_chunk$set(echo = TRUE)
```

Prerequisite: Install [SASmarkdown package](https://cran.r-project.org/web/packages/SASmarkdown/)

```{r}
ipk = function(pkg) {
  new.pkg = list.of.pkgs[!(list.of.pkgs %in% .packages(all.available = TRUE))]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE,  repos = 'http://cran.us.r-project.org')
  suppressPackageStartupMessages({
    sapply(pkg, require, character.only = TRUE)
  })
}

list.of.pkgs = c(
  "SASmarkdown", "distill"
)
ipk(list.of.pkgs)
```


## Display current `knitr` engine

```{r}
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

## Set up the SAS engine

1. You should have SAS locally installed
2. You also have the `SASmarkdown` package installed and loaded
3. Find the location of the executable SAS file. Mine is located at 
`C:\Program Files\SASHome\SASFoundation\9.4\sas.exe`
4. Set up your `knitr` options as follows, inside R chunk

```{r}
library(SASmarkdown)
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -linesize 75"
# knitr::opts_chunk$set(engine="sashtml", engine.path=saspath, 
#         engine.opts=sasopts, comment=NA)
knitr::opts_chunk$set(engine="sas", engine.path=saspath, 
        engine.opts=sasopts, comment=NA)

# run these commands to convince yourself that
# within this knitr session the engine changed.
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

## Knit to pdf or doc files

If you want to knit to pdf or doc, change the engine to `engine="sas"`. This will not use ODS output from SAS.

Following R codes are provided for reference but no executed here.

```{r, eval=FALSE}
# saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
# sasopts <- "-nosplash -linesize 75"
# knitr::opts_chunk$set(engine="sas", engine.path=saspath, 
#         engine.opts=sasopts, comment=NA)

```

## Try some `sas` code

```{sas}
proc means data=sashelp.class;
run;
```

## Even more SAS code

```{sas}
proc corr data = sashelp.class plots = matrix;
run;
```



