---
title: "Animation"
author: "bokola"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r paths, warning=FALSE, message=FALSE, results='hide'}
# Set project path

home.dir <- ifelse(Sys.info()["sysname"] == "Windows", Sys.getenv("USERPROFILE"), Sys.getenv("HOME"))
m.project.path <- path.expand(file.path(home.dir
                                        ,"Analysis"
                                        ,"TidyTuesday"
                                        ))

m.project.path <- gsub("\\\\","/",m.project.path)
if(!file.exists(m.project.path)) {
  if(dir.create(m.project.path, recursive = TRUE))
    stop("The project directory \""
         ,m.project.path
         ,"\"  has been created!\nPlease fill it with the relevant files and folders!")
  else
    stop("The project directory \""
         ,m.project.path
         ,"\"  could not be created!")
}

# Set relevant Directories

# Reporting directory
 report.location = file.path(m.project.path, "cache", "doc")
 if(!file.exists(report.location)) {
   if(dir.create(report.location, recursive = TRUE))
     stop("The project directory \""
          ,report.location
          ,"\"  has been created!\nPlease fill it with the relevant files and folders!")
   else
     stop("The project directory \""
          ,report.location
          ,"\"  could not be created!")
 }
# scripts directory
 
 scripts_location <- file.path(m.project.path, 'src', 'r')
 if(!file.exists(scripts_location)) {
   if(dir.create(scripts_location, recursive = T))
     stop('The script directory \''
          , scripts_location
          , '\' has been created!\nPlease fill it with relevant files and folders!')
   else
     stop('The script directory \''
          ,scripts_location
          ,'\' could not be created!')
 }
```

Animation is helpful when visualizing changes over time, changes over iterative/recursive techniques (like gradient descent or bellman equations), changes during a simulation.
The key point is to highlight important changes in the data over the variable that is driving the animation

```{r packages, results='hide', warning=FALSE, message=FALSE}
ipk = function(pkg){
  new.pkg = list.of.pkgs[!(list.of.pkgs %in% installed.packages()[, 'Package'])]
  if(length(new.pkg)) install.packages(new.pkg, dep = T)
  sapply(pkg, require, character.only = T)
  
}
list.of.pkgs = c("animate", "gganimate", "plotly", "googlevis", "tidyverse", "plyr", "gapminder","animation", "rust", "datasauRus", "Hmisc", "geonames", "cowplot", "ggpubs", "ggthemes")
ipk(list.of.pkgs)

```
Data

```{r data, results='hide', warning=FALSE, message=FALSE}
coast_vs_waste <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/coastal-population-vs-mismanaged-plastic.csv") %>% janitor::clean_names()
mismanaged_vs_gdp <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/per-capita-mismanaged-plastic-waste-vs-gdp-per-capita.csv" ) %>% janitor::clean_names()
waste_vs_gdp <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-21/per-capita-plastic-waste-vs-gdp-per-capita.csv") %>% janitor::clean_names() 
waste_vs_gdp = waste_vs_gdp %>% dplyr::rename(., gdp_per_capita_ppp_constant_2011_international_rate = gdp_per_capita_ppp_constant_2011_international_constant_2011_international)
plastic_data = merge(coast_vs_waste, waste_vs_gdp, all = T) %>% 
  merge(., mismanaged_vs_gdp, all = T) %>% 
  mutate(., decade = year - (year %% 10))
data = plastic_data %>% 
  ddply(., .(entity, decade), summarise,
        population = mean(total_population_gapminder))
```
First is the `animate` library
```{r, animate}
desc = c("This is a super cool example of Gradient Descent")

saveHTML({

  f1 = function(x, y) x^2 + 3 * sin(y)
  xx = grad.desc(f1, pi * c(-2, -2, 2, 2), c(-2 * pi, 2))
 
 xx$persp(col = "lightblue", theta = 30, phi = 30)

},title = "Demo of Gradient Descent", description = desc, verbose = FALSE)
```
`gganimate`
```{r gganimate}
library(datasauRus)

ggplot(datasaurus_dozen, aes(x=x, y=y))+
  geom_point()+
  theme_minimal() +
  transition_states(dataset, 3, 1) + 
  ease_aes('cubic-in-out')
```
`plotly`
```{r plotly}
library(plotly)
library(gapminder)

p <- gapminder %>%
  plot_ly(
    x = ~gdpPercap, 
    y = ~lifeExp, 
    size = ~pop, 
    color = ~continent, 
    frame = ~year, 
    text = ~country, 
    hoverinfo = "text",
    type = 'scatter',
    mode = 'markers'
  ) %>%
  layout(
    xaxis = list(
      type = "log"
    )
  )
p
```
Animated bar graphs. We use `ggplot2` and `gganimate` packages

```{r animated_bars}
library(tidyverse)
library(ggthemes)
library(animation)
library(cowplot)
library(ggpub)
library(readxl)
library(geonames)


gif_file <- "~/Dropbox/infographics/general/south_america_infant_mortality.gif"
x_space <- 0.16
speed <- 0.12
start_year <- 1900

europe <- c("Argentina", "Bolivia", "Brazil", "Chile", "Colombia", "Ecuador", "Guyana", "Paraguay", "Peru", "Suriname", "Uruguay", "Venezuela")

europe <- sort(europe, decreasing = TRUE)

data <- plastic_data %>% 
  subset(., !is.na(total))
  filter(Entity %in% europe) 
mortality %>% 
  select(Entity) %>% 
  distinct()


tweet_list <- list()
track_tweets <- 1

max(mortality$Year)
names(mortality)[4] <- "Mortality"

max_value <- max(mortality$Mortality)

breaks = seq(0,60, 10)
labels <- str_c(breaks, "%")
setwd("~/empty/")
file.remove(dir("~/empty/"))
tracker <- 1

for (yy in 1900:2015){
  yy_mortality <- mortality %>%
    filter(Year == yy) 
  
  
  
  
  gg <- yy_mortality %>%
    rename(Country = Entity) %>% 
    mutate(Country = factor(Country, levels = europe)) %>%
    ggplot() +
    geom_bar(aes(Country, Mortality), fill = "skyblue",  stat = "identity") +
    coord_flip(clip= "off") +
    theme_minimal(base_size = 6)+
    labs(x = NULL, y = NULL) +
    labs(caption = "Data source: Our World in Data", title = "The decline of infant mortality in South America")+
    labs(subtitle = "Proportion of infants dying before the age of 5")+
    geom_text(x = 13, y = max_value * 0.93, label = yy, size = 3.5) +
    theme(panel.grid.major.y = element_blank())+
    theme(panel.grid.minor.y = element_blank())+
    theme(panel.grid.minor.x = element_blank())+
    # geom_text(data = fuel_pc, aes(x = Fuel, y = fuel,  label = Percentage), hjust = 0)+
    # geom_text(data = yy_rank, aes(x = Rank, y = -0.01 * max_value, label = Fuel), hjust = 1) +
    scale_y_continuous(limits = c(0, max_value), labels = labels, breaks = breaks, expand = c(0,0)) +
    theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm")) +
    labs(fill = NULL) +
    theme(legend.position = "none")+
    theme(plot.title = element_text(face = "bold"))+
    theme(plot.margin=grid::unit(c(1,1,1,1), "mm"))
  gg
  gg_file <- str_c(str_pad(tracker, width = 3, pad = 0), ".png")
  
  ggsave(gg_file, gg, width = 8, height = 8 * 0.8, units = "cm")
  tracker <- tracker + 1
}

for (i in 1:10) {
  tracker <- tracker + 1
  gg_file <- str_c(str_pad(tracker, width = 3, pad = 0), ".png")
  
  ggsave(gg_file, gg, width = 8, height = 8 * 0.8, units = "cm")
}


# system(str_glue("convert -resize 80% -delay {100 * speed} -loop 0 *.png animation.gif"))
system(str_glue("convert -delay {100 * speed} -loop 0 *.png animation.gif"))


file.rename("animation.gif", gif_file)
```