---
title: "Understanding the complex seasonality of seasonal influenza A and B virus transmission: Evidence from six years of surveillance data in               Shanghai, China"
author: "JC paper discussion by Basil Okola"
date: "5/12/2020"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, packages, warning=FALSE,results='hide', echo=FALSE, message=FALSE}
ipk = function(pkg){
  new.pkg = list.pkg[!(list.pkg %in% installed.packages()[, "Package"])]
  if("practicalgg" %in% list.pkg){
    remotes::install_github("wilkelab/practicalgg")
  }
  if(length(new.pkg)) install.packages(new.pkg, dependencies = T)
  sapply(pkg, require, character.only = T)
}
list.pkg = c("xaringan", "xaringanthemer")
             
ipk(list.pkg)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

