---
title: "Prevalence of blindness"
author: "Basil Okola"
# abstract: |
#   **Background**: Add.
#   **Objectives**: Add.
#   **Methodology**: Add.
#    **Results**: 
#    **Conclusions**:
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    # citation_package: natbib
  # bookdown::word_document2:
  #   citation_package: natbib
  #   keep_tex: true
  # bookdown::pdf_document2:
  #   citation_package: natbib
  #   keep_tex: true
  # word_document: default
  fig_caption: yes
  tables: yes
  mainfont: "Times New Roman"
  fontsize: 11pt
  number_sections: true
  toc: no
header-includes: 
  - \usepackage{amsmath}
editor_options: 
  chunk_output_type: console
# bibliography: citation.bib
# csl: biomed-central.csl
# csl: american-political-science-association.csl
# geometry: margin=1in
# fontfamily: mathpazo
# fontsize: 12pt
# spacing: double
# bibliography: citation.bib
urlcolor: blue
link-citations: yes
linkcolor: blue
csl: "https://raw.githubusercontent.com/Bokola/resources/main/biomed-central.csl"
---


```{r setup, child="_setup.Rmd"}
# knitr::opts_chunk$set(echo = TRUE)
```

```{r data}
# set paths and packages
options(scipen = 4)
ipk = function(pkg) {
  new.pkg <-
    list.pkg[!(list.pkg %in% .packages(all.available = TRUE))]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE,
              repos = 'https://cran.us.r-project.org')
  ddpcr::quiet(sapply(pkg, require, character.only = TRUE))
}
list.pkg <- c("haven", "here", "remotes","dplyr", "purrr", "readxl", "magrittr", "janitor", "tidyr")

ipk(list.pkg)

# home  <- dirname(rstudioapi::getActiveDocumentContext()$path) %>% dirname()
# data_dir <- file.path(home, 'KI-data-coordinator', 'Data')
```

Response to the questions follow a question and answer format, first listing the question and providing an answer afterwards.


# Question 1

In calculating prevalence of blindness, we use the fact that Bilateral visual impairment is defined by the vision in the better, i.e. least impaired, eye and as such both eyes must be blind for one to be labelled blind. The prevalence of blindness is 0.02. 


```{r}
# read data
fils <- list.files(file.path(here(),  "sightsavers"),pattern = ".xlsx")
fils <- subset(fils, !grepl("~", fils))
dat <- read_xlsx(file.path(here(), "sightsavers", fils), sheet = "Data")
pop <- read_xlsx(file.path(here(), "sightsavers", fils), sheet = "pop")
pop <- pop %>% filter(age != "total")
# str(dat)
dat <- dat %>% mutate(
  blind_le = ifelse(pvale %in% c(5, 6, 7), 1, 0),
  blind_re = ifelse(pvare %in% c(5, 6, 7), 1, 0)
)

dat <- dat %>% mutate(
  blind = ifelse(blind_le == 1 & blind_re == 1, 1, 0)
)
dat <- dat %>% mutate(
  sex = ifelse(sex == 1, "male", "female")
)

tab1 <- table(dat$blind) %>% prop.table() %>% round(., 3) %>%  as.data.frame() %>%
  `colnames<-`(c("blind", "prop"))
knitr::kable(tab1, format = "pandoc")

tab2 <- dat %>% group_by(sex) %>%
  summarize(
    prop = mean(blind)
  )
knitr::kable(tab2, format = "pandoc")

p <- 0.02; pm <- 0.019;pf <- 0.024
se <- sqrt((p*(1-p))/ nrow(dat))
ci <- paste0("(", round(p - 1.96 * se, 3), ",", round(p + 1.96 * se, 3), ")")
sem <- sqrt((pm*(1-pm))/ nrow(dat))
cim <- paste0("(", round(pm - 1.96 * se, 3), ",", round(pm + 1.96 * se, 3), ")")
sef <- sqrt((pm*(1-pf))/ nrow(dat))
cif <- paste0("(", round(pf - 1.96 * se, 3), ",", round(pf + 1.96 * se, 3), ")")

```


# Question 2


The 95\% CI is given by `r ci` , for males is `r cim`, for females `r cif`


# Question 3

```{r}

dat <- dat %>%
  
mutate(
  age_g = case_when(
    age >=50 & age <= 54 ~ "50-54",
    age >=55 & age <= 59 ~ "55-59",
    age >=60 & age <= 64 ~ "60-64",
    age >=65 & age <= 69 ~ "65-69",
    age >=70 & age <= 74 ~ "70-74",
    age >=75 & age <= 80 ~ "75-80",
    TRUE ~ "80+"
  )
)

out3 <- dat %>% group_by(sex, age_g) %>%
  summarize(
    prop = round(mean(blind), 3)
  )
out3_w <- out3 %>% pivot_wider(
 names_from = sex, values_from = prop
)
names(out3_w) <- c("age", "prop_f", "prop_m")

dat3 <- out3_w %>% left_join(pop) %>% select(-total)

dat3 <- dat3 %>% mutate(
  tot_blind_male = round(male * prop_m),
  tot_blind_female = round(female * prop_f)
)

dat3$total_blind <- dat3$tot_blind_female + dat3$tot_blind_male

knitr::kable(dat3, format = "pandoc")
```

Total blind given  in last three columns

\newpage

# References
<div id="refs"></div>

# Appendix

Programs used in this report



```{r codes, eval=FALSE}
# set paths and packages
dat <- dat %>%
  
mutate(
  age_g = case_when(
    age >=50 & age <= 54 ~ "50-54",
    age >=55 & age <= 59 ~ "55-59",
    age >=60 & age <= 64 ~ "60-64",
    age >=65 & age <= 69 ~ "65-69",
    age >=70 & age <= 74 ~ "70-74",
    age >=75 & age <= 80 ~ "75-80",
    TRUE ~ "80+"
  )
)

out3 <- dat %>% group_by(sex, age_g) %>%
  summarize(
    prop = round(mean(blind), 3)
  )
out3_w <- out3 %>% pivot_wider(
 names_from = sex, values_from = prop
)
names(out3_w) <- c("age", "prop_f", "prop_m")

dat3 <- out3_w %>% left_join(pop) %>% select(-total)

dat3 <- dat3 %>% mutate(
  tot_blind_male = round(male * prop_m),
  tot_blind_female = round(female * prop_f)
)


```

```{r cod, ref.label='codes', eval=FALSE, echo=TRUE}

```


<!-- # ```{r code=readLines(knitr::purl('C:/Users/basil/Google Drive (basil.okola@student.uhasselt.be)/MSc. Stats Hasselt/y2 sem1/AMT/finite mixture project/finite-code.Rmd', documentation = 0)), eval = FALSE, echo=TRUE} -->
<!-- #  -->
<!-- # ``` -->





