---
title: "Project-Coverage of NSBDP"
author: "Basil"
date: "August 24, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(max.print = 100000)

```

This is a project to explore factors at play that impact the coverage of National School-Based Deworming Programme as implemented by the DTWI team in Kenya. The analysis is at school-level data collected during monitoring surveys by MLIS and during treatment days by respective treatment teams.

## Libraries

```{r}
ipk<-function(pkg) {
new.pkg<-list.of.pkgs[!(list.of.pkgs %in% installed.packages()[,"Package"])]
if(length(new.pkg)) install.packages(new.pkg,dependencies = T, repos = 'http://cran.us.r-project.org')
sapply(pkg,require,character.only=T)
}
list.of.pkgs<-c("foreign","dplyr","corrr","ggplot2","plotly", "tidyr","tidyverse","readstata13","magrittr","RCurl","haven","stringr","installr","shinydashboard")
ipk(list.of.pkgs)
(.packages()) #list of loaded packages
```

## Loading Data
```{r}
dat<-read.dta13("C:\\Users\\admin\\Analysis\\R\\Data\\y3_y6_vars.dta",convert.factors = F)
#dat<-read_dta(file = "https://github.com/Bokola/Analysis/raw/master/R/Data/y3_y6_vars.dta")
glimpse(dat)
dat<- dat %>% select(school_id:school_name,a_trt_total, a_reg_total,year,st_aware,st_info,sensitize_method,ct_knowledge_score,ht_knowledge_score, ct_trained,ht_trained, drugs_enough,ddm105_events_at_schl_ondd,ddm106_dd_atschl_today, worms)  %>% glimpse()
#dat %<>% select(everything()) %>% mutate(coverage=a_trt_total/a_reg_total)
dat$coverage<-dat$a_trt_total/dat$a_reg_total
cols<-c("coverage","st_info","st_aware","ct_knowledge_score","ht_knowledge_score","sensitize_method","drugs_enough","ct_trained","ht_trained","worms")
dat<-dat[complete.cases(dat[,cols]),]
cols.1<-c("coverage","st_info","st_aware","ct_knowledge_score","ht_knowledge_score","sensitize_method","drugs_enough","ct_trained","ht_trained","worms","year","county_name")
dat %<>% mutate_all(tolower)
dat%>% select(one_of(cols.1))
dat1<- dat %>% select(coverage,county_name,year,everything()) %>% rename(county=county_name) %>% glimpse() # ordering coverage first
attach(dat1) # data with vars selected for anal
dat1 %>% mutate_all(tolower)
trim <- function (x) gsub("^\\s+|\\s+$", "", x) #trimming white space
dat1$county<-trim(dat1$county)
# or
str_trim(county) # from stringr
county[grep("^.hom",as.character(county),ignore.case = T)]<-"homabay"
county[grep("^.nar",as.character(county),ignore.case = T)]<-"narok"
county[grep("^trans",as.character(county),ignore.case = T)]<-"trans nzoia" #replacing misspelled counties
a<-sort(unique(county));a
western<-county[grep("^bu|^ka|^vi",as.character(county),ignore.case = T)]
eastern<-county[grep("^ma",as.character(county),ignore.case = T)];table(eastern)
nyanza<-county[grep("^mig|^kis|^sia|^nya|^hom",as.character(county),ignore.case = T)];table(nyanza)
central<-county[grep("^kir",as.character(county),ignore.case = T)];table(central)
rift<-county[grepl("^bom|^ke|^na|^tra",as.character(county),ignore.case = T)];table(rift)
coast<-county[grep("^kil|^kwa|^lam|^momb|^tai|^tana",as.character(county),ignore.case = T)];table(coast)
dat1 %<>% mutate(region=case_when(grepl("^bu|^ka|^vi",as.character(county),ignore.case = T)~"western",grepl("^ma",as.character(county),ignore.case = T)~"eastern",grepl("^mig|^kis|^sia|^nya|^hom",as.character(county),ignore.case = T)~"nyanza",grepl("^kir",as.character(county),ignore.case = T)~"central",grepl("^bom|^ke|^na|^tra",as.character(county),ignore.case = T)~"rift",grepl("^kil|^kwa|^lam|^momb|^tai|^tana",as.character(county),ignore.case = T)~"coast"))
table(dat1$region)
rm(dat)
```
```{r cleaning strings}
# pattern matching----
a<-as.vector(county[grep("^y",as.character(county),ignore.case = T)]); unique(a) # returning 1st char in county
# to find counties with 2 or more names, search for a blank----
b<-as.vector(county[grep(" ",as.character(county), ignore.case = T)]) ; unique(b)
# counties ending with a particular char say y, use "y$"----
c<-as.vector(county[grep("y$",as.character(county), ignore.case = T)]); unique(c)
# counties whose names start by a particular letter, include the leters inside a ranged square bracket e.g "[a-k]"----
d <- as.vector(county[grep("^[a-k]",as.character(county),ignore.case = T)]); unique(d)
# use negative subscripts if not interested in a particular pattern----
e <- as.vector(county[-grep("[a-d]$",as.character(county), ignore.case = T)]); unique(e)
# looking for the ith char, use {} ----
f <- as.vector(county[grep("^.{2}[a-k]",as.character(county),ignore.case = T)]); unique(f) # 3rd char within a-k
g <- as.vector(county[grep("^.[a-k]",as.character(county),ignore.case = T)]); unique(g) # 2nd char within a-k
# returning lengths ----
h <- as.vector(county[grep("^.{,5}$",as.character(county),ignore.case = T)]); unique(h) # up to 5 chars
i <- as.vector(county[grep("^.{5,}$",as.character(county),ignore.case = T)]); unique(i) # at least 5 chars
k <- as.vector(county[grep("^trans",as.character(county),ignore.case = T)]); unique(k)
l <- as.vector(county[grep("k$",as.character(county),ignore.case = T)]); unique(l)
```
```{r strings-to-factors}
get.label(dat1,"st_info")
dat1$st_info<-factor(dat1$st_info,levels=c('1','2','3'),labels = c('Deworming info 1: Assembly Announcement/teachers','Deworming info 2: Other students/friends','Deworming info 3: Parents'))
get.label(dat1,"worms")
dat1$worms<-factor(dat1$worms,levels = c('0','1'),labels = c('STH/Soil Transmitted',' SCHISTO/Bilharzia'))
get.label(dat1,'sensitize_method')
dat1$sensitize_method<-factor(dat1$sensitize_method,levels = c('1','2','3','4','5','6'),labels = c('Conduct Health Education in Class','Display Posters in the School','Discuss Deworming Day at School Management Meetings','Conduct ECD Outreach','Encourage Children to share Deworming Day Information with Parents','None'))
dat1$drugs_enough<-factor(dat1$drugs_enough,levels = c('0','1'),labels = c('Drugs enough: No','Drugs enough: Yes'))
dat1$year<-factor(year,levels=c("3","4","5","6"),labels = c("year 3","year 4","year 5","year 6"))
num<-c("st_aware","ct_knowledge_score","ht_knowledge_score","coverage")
dat1[,num]<-lapply(dat1[,num], as.numeric)

glimpse(dat1)
```

## Exploration
```{r}
p1<-ggplot(dat1,aes(coverage))+geom_histogram(aes(y=..density..),binwidth = 0.01)+geom_density()+ggtitle("Distribution of NSBDP reported coverages")
p1
#Scatter
scatter_by<- function(data,...) {
  ggplot(data)+geom_point(aes(...))
}
p2<-scatter_by(dat,factor(sensitize_method),coverage,shape=sensitize_method) +scale_x_discrete(labels=c("1","2","3","4","5","6"))

boxplot_by<-function(data,...) {
ggplot(data)+geom_boxplot(aes(...))
}
boxplot_by(dat,sensitize_method,coverage,outlier.color="Red",color=sensitize_method)+scale_x_discrete(labels=c("1","2","3","4","5","6"))
boxplot_by(dat,worms,coverage,outlier.color="Black",color=worms,na.rm=T)+scale_x_discrete(labels = c("0","1"))
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
