---
title: "Project"
author: "Basil"
date: "August 24, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Notes
This is a project to explore factors at play that impact the coverage of National School-Based Deworming Programme as implemented by the DTWI team in Kenya. The analysis is at school-level data collected during monitoring surveys by MLIS and during treatment days by respective treatment teams.

## Libraries

```{r}
ipk<-function(pkg) {
new.pkg<-list.of.pkgs[!(list.of.pkgs %in% installed.packages()[,"Package"])]
if(length(new.pkg)) install.packages(new.pkg,dependencies = T, repos = 'http://cran.us.r-project.org')
sapply(pkg,require,character.only=T)
}
list.of.pkgs<-c("foreign","dplyr","corrr","ggplot2","plotly", "tidyr","readstata13","magrittr","RCurl","haven","stringr")
ipk(list.of.pkgs)
(.packages()) #list of loaded packages
```

## Loading Data
```{r}
dat<-read.dta13("C:\\Users\\admin\\Analysis\\R\\Data\\Small_holders_farmers_survey_Clean Data_ 6 Aug 2018.dta",convert.factors = F)
#dat<-read_dta(file = "https://github.com/Bokola/Analysis/raw/master/R/Data/y3_y6_vars.dta")
glimpse(dat)
names(dat)
dat<-dat[,order(names(dat))] # oerering variable names
glimpse(dat)
View(dat)
dat<- dat %>% select(school_id:school_name,a_trt_total, a_reg_total,year,st_aware,st_info,sensitize_method,ct_knowledge_score,ht_knowledge_score, ct_trained,ht_trained, drugs_enough,ddm105_events_at_schl_ondd,ddm106_dd_atschl_today, worms) 
glimpse(dat) %>% attach() 
#dat %<>% mutate(coverage=a_trt_total/a_reg_total)
dat$coverage<-a_trt_total/a_reg_total
cols<-c("coverage","st_info","st_aware","ct_knowledge_score","ht_knowledge_score","sensitize_method","drugs_enough","ct_trained","ht_trained","worms")
dat<-dat[complete.cases(dat[,cols]),]
glimpse(dat)
attach(dat)
#get.origin.codes(st_info,label.table) #recreate code numbers of a factor as stored in stata data set --x must be a factor
# converting to factors
get.label(dat,"st_info")
dat$st_info<-factor(dat$st_info,levels=c('1','2','3'),labels = c('Deworming info 1: Assembly Announcement/teachers','Deworming info 2: Other students/friends','Deworming info 3: Parents'))
get.label(dat,"worms")
dat$worms<-factor(dat$worms,levels = c('0','1'),labels = c('STH/Soil Transmitted',' SCHISTO/Bilharzia'))
get.label(dat,'sensitize_method')
dat$sensitize_method<-factor(dat$sensitize_method,levels = c('1','2','3','4','5','6'),labels = c('Conduct Health Education in Class','Display Posters in the School','Discuss Deworming Day at School Management Meetings','Conduct ECD Outreach','Encourage Children to share Deworming Day Information with Parents','None'))
dat$drugs_enough<-factor(dat$drugs_enough,levels = c('0','1'),labels = c('Drugs enough: No','Drugs enough: Yes'))
dat$year<-factor(year,levels=c("3","4","5","6"),labels = c("year 3","year 4","year 5","year 6"))

```

## Exploration
```{r}
p1<-ggplot(dat,aes(coverage))+geom_histogram(aes(y=..density..))+geom_density()+ggtitle("Distribution of NSBDP reported coverages")
#Scatter
scatter_by<- function(data,...) {
  ggplot(data)+geom_point(aes(...))
}
p2<-scatter_by(dat,factor(sensitize_method),coverage,shape=sensitize_method) +scale_x_discrete(labels=c("1","2","3","4","5","6"))

boxplot_by<-function(data,...) {
ggplot(data)+geom_boxplot(aes(...))
}
boxplot_by(dat,sensitize_method,coverage,outlier.color="Red",color=sensitize_method)+scale_x_discrete(labels=c("1","2","3","4","5","6"))
boxplot_by(dat,worms,coverage,outlier.color="Black",color=worms,na.rm=T)+scale_x_discrete(labels = c("0","1"))
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
