---
title: "shiny"
author: "Owiti Okola"
date: "4 September 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R shiny
Goal is to learn shiny and build apps, incorporating ggplot2 and dplyr. Shiny objects are contained in single scripts called app.r which can be run by runApp("newdir")
app.r has three components; a user interface object, a server function and a call to the shinyApp function
The user interface(ui) object controls layout and appearance of the app, server function the instructions to build the app and shinyApp function creates shiny app objects from explicit UI UI/serv
UI/server pair.

```{r packages}
ipk<-function(pkg) {
new.pkg<-list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.pkg)) install.packages(new.pkg,repos = "https://cran.us.r-project.org",dependencies = T)
sapply(pkg,require,character.only = T)
}
list.of.packages<-c("dplyr","ggplot2","shiny")
ipk(list.of.packages)
```
## Inside shiny
```{r}
runExample("01_hello")
# ui
ui<-fl

```