---
title: "R Notebook"
output: html_notebook
---
## Packagess


```{r}
ipk <- function(pkg) {
  new.pkgs <- list.of.pkgs[!(list.of.pkgs %in% installed.packages()["Package"])]
  if(length(new.pkgs)) install.packages(new.pkgs, repos = 'https://cran.us.r-project.org', dependencies = T)
  sapply(pkg, require, character.only = T)
}
list.of.pkgs <- c("shiny", "tidyverse")
ipk(list.of.pkgs)
```
## ui

```{r}
ui <- fluidPage(
  titlePanel("shiny app"),
  sidebarLayout(
    sidebarPanel(
      sliderInput(inputId = "bins", label = "number of bins", min = 1, max = 50, value = 30)
    ),
    mainPanel(
      plotOutput(outputId = "distPlot")
    )
    )
  )
  

```
## Server

```{r}
server <- function(input, output) {
  
  
  
  
  output$distPlot <- renderPlot ({
    
    x <- faithful$waiting
    bins <- seq(min(x), max(x), length.out = input$bins + 1)
    hist(x, breaks = bins, col = "dark grey", border = "white",
         xlab = "Waiting time to next eruption (in mins)",
         main = "Histogram of waiting time")
    })
}

```
## calling the app
```{r}
shinyApp(ui = ui, server = server)

```
Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
