---
title: "data.table"
author: "Owiti Okola"
date: "October 16, 2018"
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    highlight: tango
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '3'
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
knitr::opts_knit$set(normalizePath("C:\\Users\\admin\\Google Drive\\Pgd\\Redcap exercises ---- interview"))
options(max.print = 100)

```
# 1) **Inside the `data.table` Package and it's functionality in data analysis**
This Markdown document illustrates the functionality of `data.table` in data wrangling tasks. The primary goal of `datatable` is to reduce programming time (fewer function calls, less variable name repetition) and the second goal is to reduce compute time.
First we remove existing objects using `rm(list = ls())` then clear the work space using the `cat("\014")` function and later set up working directory, and install the `data.table` package. An object in `data.table` takes the form `A[i,j,by()]`; `i` maps to `WHERE` in SQL, `j` maps to `SELECT` in SQL and `by` maps to `GROUP BY` in SQL. The general form is "Take DT, subset rows using `i', then calculate `j' grouped `by`".
```{r}
rm(list=ls()) # clears objects
cat("\014") # clears console
#setwd("C:\\Users\\admin\\Google Drive\\Pgd\\Redcap exercises ---- interview")
getwd()
ipk <- function(pkg) {
  new.pkg <- list.of.pkg[!(list.of.pkg %in% installed.packages()[, "Package"])]
  if(length(new.pkg)) install.packages(new.pkg, repos = 'https://cran.us.r-project.org', dependencies = T)
  sapply(pkg, require, character.only = T)
}
list.of.pkg <- c("tidyverse", "data.table", "magrittr")
ipk(list.of.pkg)
```

We can use **data.table**'s fast file reader `fread` to load flights directly as follows: `fread`  also accepts `http` and `https` URLs directly as well as operating system commands such as `sed` and `awk` output.
```{r}
setwd("C:\\Users\\admin\\Google Drive\\Pgd\\Redcap exercises ---- interview")
data <- fread("SampleData.csv")
glimpse(data)
data %<>% rename(serial = V1) %>% mutate_all(tolower) %>% as.data.table()
data[,date_adm:=parse_date(date_adm, format = "%m/%d/%Y")]
data[,date_discharge:=parse_date(date_discharge, format = "%m/%d/%Y")]
data[, age_days:=as.numeric(age_days)]
data[, age_mths:=as.numeric(age_mths)]
data[,age_years:=as.numeric(age_years)]
class(data)
glimpse(data)
```
# **Basics **
## a) **What is a data.table?**
`data.table` is an R package that provides an enhanced version of *data.frames* . in the previous chunk we created a *data.table* using `fread()`. We can also create a *data.table* using `data.table()` or coerse objects to *data.table* using `as.data.table()`

Note that:

  * Unlike *data.frames*, columns of `character` type are never converted to             `factors` by default.
  
  * Row numbers are printed with a `:` in order to visually separate the row number      from the first column.
  
  * When the number of rows to print exceeds the global option                          `datatable.print.nrows` (default = 100), it automatically prints only the top 5       and bottom 5 rows.
```{r}
getOption("datatable.print.nrows") # 100
```
 * *data.table* doesn't set or use row names, ever. We will see as to why in "*Keys      and fast binary search based subset*" vignette.

##  b) **General form of data.table - in what way is data.table enhanced?**
In contrast to *data.frame*, you can do *a lot more* than just subsetting rows and selecting colums within the frame of a *data.table*, i.e., within `[ ... ]`. To understand it we will have to first look at the general form of a *data.tale* syntax, as shown below:
````{r}
# DT[i, j, by]
## R:   i       j               by
## SQL: where   select|update   group by
```

The way to read it (out loud) is: Take `DT`, subset rows using `i`, then calculate `j`, grouped `by`.

## c) **Subset rows in `i`**

Get all the rows where `hosp_id` is "Kakamega provincial general hospital"
```{r}
ans <- data[grepl("^kaka", as.character(hosp_id), ignore.case = T),]
head(ans)
```
* Within the frame of a *data.table*, columns can be referred to as if they are       variables. Therefore we simply refer to `hosp_id` as if it is a variable. We do     not need the prefix `data$` each time. However using `data$hosp_id` would work      just fine.
* The row indices that satisfies the condition `hosp_id == grepl("^kaka")` are         computed and since there is nothing else to be done, a *data.table* of all columns   from `data` corresponding to to those row indices are returned.
* A comma after the condition is also not required in `i`. But `data[hosp_id ==       grepl("^kaka"),]` would work just fine. In *data.frames* however, the comma is       necessary.

Get the first two rows from `data`
```{r}
ans <- data[1:2]
head(ans)
```
In this case, there is no condition. The row indices are already provided in `i`. We therefore return a `data.table` with all columns corresponding to the row indices.

Sort `data` by `id` column in ascending order, and then by `hosp_id` in descending order
```{r}
ans <- data[order(id, -hosp_id)]
head(ans)
```
### **`order()` is internally optimised **

* We can use "-" on a character column within the frame of a *data.table* to sort in   decreasing order.

## d) **Select columns in `j` **

Select unique entries of `hosp_id` column, but return it as a vector
```{r}
ans <- data[,unique(hosp_id)]
head(ans)
ans <- data[,hosp_id]
head(ans)
```
Select `hosp_id` column and return as a *data.table* instead
```{r}
ans <- data[, list(hosp_id)]
head(ans)
typeof(ans)
class(ans)
```
* We wrap the variables within `list()`, which ensures that a *data.table* is         returned. In case of a single column name, not wrapping with `list()` returns a     vector instead, as seen previously.
*  *data.table* also allows using `.()` to wrap columns with. It is an *alias* to     `list()`. You can use whichever you prefer. We'll be using `.()` henceforth.
  *data.tables* (and *data.frames*) are internally lists as well, but with all its columns with equal length and with a *class* attribute. Allowing `j` to return a *list* enables converting and returning a *data.table* more efficient.

**Tip**:
As long as `j-expression` returns a *list*, each element of the list will be converted to a column in the resulting *data.table*.This makes `j` quite powerful.

* - selecting both `id` and `hosp_id` columns
```{r}
ans <- data[,.(id, hosp_id)]
head(ans,n=10)
## Alternatively
ans <- data[,list(id, hosp_id)]
head(ans)
```
Wrap both columns in `.()` or `list()`. That is it
* - Selecting both `id` and `hosp_id` cols and renaming them to `patient_id` and `hosp_name`

```{r}
ans <- data[,.(patient_id = id, hosp_name = hosp_id)]
head(ans)
```


## e) **Compute or do in `j`**

We calculate mean age in months:

We first create the new variable `age_in_mths`
```{r}
data[, age_in_mths:= ifelse(is.na(age_days), 0, age_days/30) + ifelse(is.na(age_mths), 0, age_mths) + ifelse(is.na(age_years), 0, age_years *12)]

ans <- data[, .(mean_age = mean(age_in_mths))]
head(ans)
```

How many admissions were done at kakamega general in the month of june?
```{r}
ans <- data[grepl("^kaka", ignore.case =  T, as.character(hosp_id)) & month(date_adm) == 6L, length(id)]
head(ans)
```
Note that we did not wrap `.N` with a `list()` or `.()`. Therefore a vector is returned.
We could have accomplished the same operation by doing `nrow(data[grepl("^kaka", as.character(hosp_id), ignore.case = T) & month(date_adm) == 6)])` which is unnecessary and inefficient.

This kind of operation occurs very frequently, especially while grouping. The *data.table* package provides a special symbol `.N` for it.

```{r}
ans <- data[grepl("^kaka", as.character(hosp_id), ignore.case = T) & month(date_adm) == 6, .N]
head(ans)
```

Note that we did not wrap `.N` with a `list()` or `.()`. Therefore a vector is returned.

## f) **Subset in `i` and do in `j`**

What was the mean admission for kakamega general in the month of june?

```{r}
ans <- data[grepl("^kaka", ignore.case = T, as.character(hosp_id)) & month(date_adm) == 6L, .(.N, mean_adm =.N/nrow(data) *100)]
head(ans)
```

## g) **Referring to columns by names in `j` like in *data.frame* **

If you are writing out the column names explicitly, there is no difference vis-a-vis `data.frame`


* Select columns the `data.frame` way


We can select both `id`, `hosp_id` and `date_adm` like this

```{r}
ans <- data[, c("id", "hosp_id", "date_adm")]
head(ans)
```


If you have stored the desired columns in a character vector, there are two options: the `..` prefix, or using the `with` argument

* Select columns named in a variable using the `..` prefix

```{r}
select_cols <- c("id", "hosp_id", "date_adm")
data[, ..select_cols]
```

We can also deselect columns using `-` or `!` e.g.

```{r}
ans <- data[, -c("id", "hosp_id", "date_adm")]
head(ans)
ans <- data[, !c("id", "hosp_id", "date_adm")]
head(ans)
```
We can also select by specifying `start:end` 

```{r}
ans <- data[, id:date_discharge]
head(ans)
```


* Selecting columns using `with()`

We select entries for which ages were greater than >= 24 months
```{r}
ans <- data[with(data, age_in_mths >= 24)]
nrow(ans)
```

# 2) **Aggregation**

## a) **Grouping using `by`**

In this section we explore how `i`, `j`, can be combined with `by` to perform operations *by group*

* How do find number of admissions by `hosp_id` or even  `sex`?

```{r}
ans <- data[, .(.N), by = .(hosp_id)]
top_n(ans, -5)
ans[order(-N, hosp_id)]
```
 *Since we did not provide a name for the column returned in  `j`, it was named `N`   automatically by recognising the special symbol `.N`
 *When there is only one column or expression to refer to in `j` and `by`, we can    drop the `.()` notation. This is purely for convenience
```{r}
ans <- data[, .N, by = child_sex]
head(ans)
ans <- data[,.(count =.N), by =.(hosp_id)] # allows renaming of cols-- my fav
ans[order(-count, hosp_id)]
```
What if we just want to count admissions in Kakamega hosp?
```{r}
ans <- data[grepl("^kaka", as.character(hosp_id), ignore.case = T), .(count = .N)]
ans
```
What if we want to group by more than one column? You just supply the respective col names within `by`
```{r}
ans <- data[,.(count = .N), by = .(hosp_id, child_sex)]
ans[order(hosp_id,child_sex, -count)]
```

Now what if we would like to order the results by those grouping columns?

## b) **keyby**
*data.table* retaining the original order is intentional and by design. There are cases when preserving the original order is essential. But at times we would like to automatically sort by the variables we grouped by by supplying `keyby` in  place of `by`:
```{r}
ans <- data[, .(count = .N), keyby = .(hosp_id, child_sex)]
ans
```
## c) **Chaining**

Consider when you have to group and order at the same time. Chaining comes in handy. We can tack expressions one after another, forming a chain of operations, i.e., `DT[ ... ][ ... ][ ... ]`

```{r}
ans <- data[, .(count = .N), by = .(hosp_id, child_sex)][order(hosp_id, - count)]
head(ans)
```

## d) **Expressions in `by`**
`by` does not just take column names but expression as well
```{r}
ans <- data[, .(count = .N), .(female = grepl("^fe", as.character(child_sex), ignore.case = T), aged_24mo_andover = age_in_mths >= 24)][order(-female, - aged_24mo_andover, count)]
ans
```

## e) **Multiple columns in `j` - `.SD` **

* Do we have to compute `mean()` for each column individually? it's impractical to have to type `mean(myclol)` for every column one by one especially when you have columns running into their hundreds or even millions. What next? How can we do this efficiently? Going back to the tip - "As long as j-expression returns a list, each element of the list will be converted to a column in the resulting *data.table*". Suppose we can refer to the data subset for each group as a variable while grouping, then we can loop through all the columns of that variable using the already familiar base function `lapply()`. We don't have to klearn any new function.

### **Special symbol `.SD` **
*data.table* provides a special symbol called `.SD`. It stands for Subset of Data. It by itself is a *data.table* that holds the data for the *current group* defined using `by`. 

Recall that a *data.table* is internally a list as well with all its columns of equal length

```{r}
DT = data.table(ID = c("b","b","b","a","a","c"), a = 1:6, b = 7:12, c = 13:18)
DT
DT[, print(.SD), by = ID]
DT[,lapply(.SD, mean), by = ID]
```

* `.SD` holds the rows corresponding to columns a, b and c for that group. We         compute the `mean()` on each of these columns using the already familiar base       function `lapply()`.

* Each group returns a list of three elements containing the mean value which will 
  become the columns of the resulting `data.table`.
  
* Since `lapply()` returns a *list*, there is no need to wrap it with an additional   `()` 
So what if you only want to apply function to specific columns. `.SD` would contain all the columns other than the grouping variables by default.

### **`.SDcols` **

We use the `.SDcols` argument. It accepts column names or column indices. For example, `.SDcols = c("age_mths", "age_in_mths", "age_days", "age_years")` ensures that `.SD` conntains only these four columns for each group. You can also provide columns to remove using `!` and `-`

Lets use both `.SD` and `.SDcols`
```{r}
data[grepl("^kaka", as.character(hosp_id), ignore.case = T), ## only for kakamega 
     lapply(.SD, mean), ## compute the mean
     by = .(child_sex, convulsions), ## for every gender and convulsion state
     .SDcols = c("age_in_mths", "age_years")] # for just those specified in .SDcols
```

## f) **Subset `.SD` for each group: **
How can we return the first two rows of each year of admission?
```{r}
ans <- data[, head(.SD, 2), by = year(date_adm)][order(year)]
head(ans)
```
* `.SD` is a *data.table* that holds all the rows for that *group*. We simply subset the first two rows as we have seen here already.

* For each group, `head(.SD, 2)` returns the first two rows as a *data.table* whic is also a list. So we do not have to wrap it with `.()`.



