---
title: "A dive into dplyr"
author: "Basil"
date: "October 13, 2018"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    highlight: tango
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: '3'
---
We use the tidyverse package to demonztrate it's usefulness in data wrangling.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```
First we install required packages

```{r, echo= TRUE, include=TRUE}
ipk <- function (pkg) {
  new.pkg <- list.of.pkgs [!(list.of.pkgs %in% installed.packages()[,"Package"])]
  if(length(new.pkg)) install.packages(new.pkg,repos = 'https://cran.us.r-project.org',dependencies = T)
  sapply(pkg, require, character.only = T)
}
list.of.pkgs <- c("tidyverse","magrittr")

ipk(list.of.pkgs)
```

# **Data Wrangling Part 1: Basic to Advanced Ways to Select Columns** 

Covers ordering, naming and selecting columns: `select_all()`, `select_if()` and some shortcuts like `everything()`

```{r}
# built-in R dataset
glimpse(msleep)
```

## **Selecting columns**
### **Selecting columns: The basics**

The order in which you add them determines the order in which they appear on the output
```{r}
msleep %>% select(name, genus, sleep_total, awake) %>% glimpse()
```
You can use chunks: `start_col:end_col` syntax or even `deselect columns` by adding a minus sign in front of the column names
```{r}
msleep %>% select(name:order, sleep_total:sleep_cycle) %>% glimpse()
msleep %>% select(-conservation,-(sleep_total:awake)) %>% glimpse()
```
It's possible deselect a whole chunk, and then re-add a column again
```{r}
msleep %>% select(-(name:awake), conservation) %>% glimpse
```
 ## **Selecting columns based on partial column names** 
 
Entails using `starts_with()`, `ends_with()` or `contains()`
```{r}
msleep %>% select(name, starts_with("sleep")) %>% glimpse
msleep %>% select(contains("eep"), ends_with("wt")) %>% glimpse
```
## **Selecting columns based on regex**

The previous helper functions work with exact pattern matches. If you have similar patterns that are not entirely the same, you can use any regular expression inside `matches()`. Code below adds columns that contain an 'o' followed by one or more other letters, and 'er'.

```{r}
msleep %>% select(matches("o.+er")) %>% glimpse
```
## **Selecting  columns based on pre-identified columns**

You can specify column names in a vector and use `one_of()` function within `select()` or using the !! operator to select them
```{r}
col <- c("name", "genus","vore", "order", "conservation")
msleep %>% select(one_of(col)) %>% glimpse
msleep %>% select(!!col) %>% glimpse
```

## **Selecting columns based on their data types**

The `select_if()` function allows you to pass functions which return logical statements. For instance you can select all string columns using `select_if(is.character)` . Similarly, you can add `is.numeric`, `is.integer`, `is.double`, `is.logic`, `is.factor` .

```{r}
msleep %>% select_if(is.numeric) %>% glimpse
```
You can also select the negation but must add a tilde to ensure you still pass a function to `select_if` . The select_all/if/at functions require that a function is passed as an argument. If you have to add negation or arguments, you will have to wrap your functions inside `funs()` or add a tilde before to remake it a function.
```{r}
msleep %>% select_if(~!is.numeric(.)) %>% glimpse
```
## **Selecting columns by logical expressions**

`select_if()` allows you to select based on any logical function, not just the data types. It is possible to select all columns with an average above 500 for instance, but you'd have to select only numeric columns upfront to avoid errors. Similarly, mean > 500 is not a function in itself, so you need to a tilde upfront or wrap inside the `funs()` to turn the statement into a function.

```{r}
msleep %>% select_if(funs(!is.character(.))) %>% select_if(funs(mean(., na.rm = T)>10))
```
Or shorter:
```{r}
msleep %>% select_if(funs(!is.character(.) & mean(., na.rm = T)>10))
```
or even shorter
```{r}
msleep %>% select_if(~is.numeric(.) & mean(., na.rm = T)> 10)
```
Another useful function for `select_if` is `n_distinct()`, which counts the amount of distinct values that can be found in a column
```{r}
msleep %>% select_if(funs(n_distinct(.)<10))
#or
msleep %>% select_if(~n_distinct(.)<10)
```

## **Re-ordering columns**

with `select()`, the order in which you add the columns determine their final order
```{r}
msleep %>% select(conservation, contains("eep"), name) %>% glimpse
```

if you want to move a few columns to the front, use  `everything()` afterwards which will add all the columns and save you typing

```{r}
msleep %>% select(conservation, sleep_total, everything()) %>% glimpse
```
## **Column names**
Sometimes all you want to change are the column names

### **Renaming columns**
You can directly rename within the `select()` function
```{r}
msleep %>% select(animal = name, sleep_total, extinction_threat = conservation) %>% glimpse
```
If you want to retainn all columns and thus do not require the `select()` function, you can rename by adding a `rename()` statement

```{r}
msleep %>% rename(animal = name, extinction_threat = conservation) %>% glimpse
```
### **Reformatting all column names**
The `select_all()` function allows changes to all columns, and takes a function as an argument.
Format cases using `toupper()` and `tolower()` functions
```{r}
msleep %>% select_all(toupper)
msleep %>% select_all(tolower)
```
If you have messy variable names especially those coming from excel, you can replace all white spaces with an underscore for instance
```{r}
# making an unclean database
msleep2 <- select(msleep, name, sleep_total, brainwt)
colnames(msleep2) <- c("name", "sleep total", "brain wight")
msleep2 %>% select_all(~str_replace(., " ", "_"))
```
or in case your ccolumns contain other meta-data like question numbers
```{r}
# making an unclean database 
msleep2 <- select(msleep, name, sleep_total, brainwt)
colnames(msleep2) <- c("Q1 name", "Q2 sleep total", "Q3 brain weight")
msleep2[1:3,]
```
You can use `select_all()` in combination with `str_replace` to get rid of the extra characters
```{r}
msleep2 %>% select_all(~str_replace(., "Q[0-9]+", "")) %>% select_all(~str_replace(., " ","_"))
```

### **Row names to column **
Some dataframes have rownames that are not actually a column in itself, like the mtcars dataset:
```{r}
mtcars %>% head
```
If you want this column to be an actual column, you can use the `rownames_to_column()` function and specify a new column name.
```{r}
mtcars %>% tibble::rownames_to_column("car_model") %>% head
```

# **Data Wrangling Part 2: Transforming your columns into the right shape **
This section covers tools to manipulate your columns to get them the way you want them; includes calculation of new columns, changing a column unto discrete values or splitting/merging columns.

## **Mutating columns: The basics **
You can make new columns or modify existing ones with `mutate()` function. One of the simplest options is calculations based on values of other columns.
```{r}
msleep %>% select(name, sleep_total) %>% mutate(sleep_total_min = sleep_total *60)
```
New columns can be created using aggregate functions such as average, median, max, min, sd e.t.c.
```{r}
msleep %>% select(name, sleep_total) %>% mutate(sleep_total_vs_AVG = sleep_total - round(mean(sleep_total),1), sleep_total_vs_MIN = sleep_total - min(sleep_total))
```
If you want to `sum()` or `mean()` across columns , you might run into errors or absurdd answers.In such cases you may resort to spelling out the arithmetics: `mutate(average = (sleep_rem + sleep_cycle)/2)` or adding a special instruction to the pipe that it should perform those aggregate functions not on the entire column, but by row:
```{r}
# alternative to using the actual arithmetics
msleep %>% select(name, contains("sleep")) %>% rowwise() %>% mutate(avg = mean(c(sleep_rem, sleep_cycle)))
```
The `ifelse()` function enables you specify a logical statement detailing what happens if the statement returns `TRUE` or `FALSE`.
```{r}
msleep %>% select(name, brainwt) %>% mutate(brainwt2 = ifelse(brainwt > 4, NA, brainwt)) %>% arrange(desc(is.na(brainwt))) %>% glimpse
```
You can also mutate string columns with stringr's `str_extract()` function in combination with any character or regex patterns. Code below returns the last word of the animal name and makes it lower case.
```{r}
msleep %>% select(name) %>% mutate(name_last_word = tolower(str_extract(name, pattern = "\\w+$"))) 
```
## **Mutating several columns at once **
As with `select()` there are variants to `mutate()`: 
* `mutate_all()` will mutate all columns based on supplied columns
* `mutate_if()` requires a function that returns a boolean to select columns. If that is true, the mutate instructions will be followed on those variables.
* `mutate_at()` requires you to specify columns inside a `vars()` argument for which the mutation will be done.
### **Mutate all **
```{r}
msleep %>% mutate_all(tolower)
```
The mutating action needs to be a function: in many cases you can pass the function name without the brackets. you can make a function upfront or on the fly with the tilde. When making a function on the fly, you need a way to refer to the value you are replacing: which is what the `.` symbolizes.

```{r}
msleep_ohno <- msleep %>% mutate_all(~paste(., " /n "))
msleep_ohno[,1:4]
```
To clean this up, first remove any `/n`, and then trim additional white spaces.
```{r}
msleep_corr <- msleep_ohno %>% mutate_all(~str_replace_all(., "/n", "")) %>% mutate_all(str_trim)
msleep_corr[,1:4]
```
### **Mutate if **
By using `mutate_if()` we need two arguments inside a pipe:
 * First is information about the columns you want to consider: Easiest functions include `is.numeric`, `is.integer`, `is.double`, `is.logical`, `is.factor`, `bubridate::is.POSIXt` or `lubridate::is.Date`.
 * Second is the instructions about the mutation in form of a function. If needed use a tilde or `funs() before.
```{r}
msleep %>% select(name, sleep_total:bodywt) %>% mutate_if(is.numeric, round)
```

### **Mutate at to change specific columns **
By using `mutate_at()` we need two arguments inside a pipe:
* First it needs information about the columns you want to consider. In this case you can wrap any selection of columns using all the options available inside a `select()` function and wrap it inside `vars()`.
* Secondly it needs instructions about the mutation in the form of a function. If needed, use ilde or `funs()` before
```{r}
msleep %>% select(name, sleep_total:awake) %>% mutate_at(vars(contains("sleep")), ~(.*60))
```
## **Changing column names after mutation **
```{r}
msleep %>% select(name, sleep_total:awake) %>% mutate_at(vars(contains("sleep")), ~(.*60)) %>% rename_at(vars(contains("sleep")), ~paste0(.,"_min"))
```
Alternatively, you can assign a "tag" inside `funs()` which will be appended to the current name.The difference is that `funs()` version is one code less, but columns will be added rather than replaced.
```{r}
msleep %>% select(name, sleep_total:awake) %>% mutate_at(vars(contains("sleep")), funs(min = .*60))
```
## **Working with discrete columns **
### **Recoding discrete columns**
To rename or reorganize current discrete columns, you can use `recode()` inside a `mutate()` statement: this enables you to change the current naming, or to group current levels to less levels. The `.default` refers to anything that isn't covered by the before groups with the exception of NA. You can change NA into something other than NA by adding a `.missing` argument if you want.
```{r}
msleep %>% mutate(conservation2 = recode(conservation, "en" = "Endangered", "lc" = "Least_Concern", "domesticated" = "Least_Concern", .default = "other")) %>% count(conservation2)
```
A special version exists to return a factor: `recode_factor()`. By default the `.ordered` argument is `FALSE`. 
```{r}
msleep %>% mutate(conservation2 =recode_factor(conservation, "en" = "Endangered","lc" = "Least_Concer", "domesticated" = "Least_Concern", .default = "other", .missing = "no data", .ordered = T)) %>% count(conservation2) %>% ggplot() + geom_col(mapping = aes(x = conservation2, y = n))
```


## **Creating new discrete column (two levels)**
The `ifelse` statement can be used to turn a numeric column to turna numeric column into a discrete one (with two levels)
```{r}
msleep %>% select(name, sleep_total) %>% mutate(sleep_time = ifelse(sleep_total > 10, "long", "short"))
```
## **Creating new discrete column (multiple levels) **
`case_when()` allows as many statements as you want.
For everything that is left at the end just use `TRUE ~ "newname"`. Unfortuantely there is no easy way to get `case_when()` to return an ordered factor, you can use `forcasts::fct_level()`, or just with a `factor()` function.
```{r}
msleep %>% select(name, sleep_total) %>% mutate(sleep_total_discr = case_when(sleep_total > 13 ~ "very long", sleep_total > 10 ~ "long", sleep_total > 7 ~ "limited", TRUE ~ "short")) %>% mutate(sleep_total_discr = factor(sleep_total_discr, levels = c("short", "limited", "long", "very long")))
```
The `case_when()` function does not only work inside a column, but can be used for grouping across columns:
```{r}
msleep %>% mutate(silly_groups = case_when(brainwt < 0.001 ~ "light headed", sleep_total > 10 ~ "lazy_sleeper", is.na(sleep_rem) ~ "absent_rem", TRUE ~ "other")) %>% count(silly_groups)
```
## **Splitting and merging columns**
We use [this](C:\\Users\\Basil MLIS\\Analysis\\R\\Data\\conservation_explanation.csv) dataset for illustrations
```{r}
#(conservation_expl <- read_csv("C:\\Users\\Basil MLIS\\Analysis\\R\\Data\\conservation_explanation.csv")) 
(conservation_expl <- tibble(`conservation abbreviation` = c("EX = Extinct", "EW = Extinct in the wild", "CR = Critically Endangered", "EN = Endangered", "VU = Vulnerable", "NT = Near Threatened", "LC = Least Concern", " DD = Data deficient", " NE = Not evaluated", " PE = Probably extinct (informal)", "PEW = Probably extinct in the wild (informal)")))
(conservation_expl %<>% rename_all(
  funs(
    stringr::str_replace_all(., " ","_") %>% 
      stringr::str_to_lower(.)
    )
  )
)
```
You can unmerge any columns by using tidyr's `separate()` function. You have to specify the column to be splitted followed by the new column names, and the separator to be looked for.
```{r}
(conservation_tab <- conservation_expl %>% separate(conservation_abbreviation, into = c("abbreviation", "description"), sep = "="))
```
The opposite is tidyr's `unite()` function. You specify the new column name, and then  the columns to be united, and the separator you intend to use.
```{r}
conservation_tab %>% unite(united_col, abbreviation, description, sep = ":")
```
## **Bringing in columns from other data tables **
You can use joining functions from `dplyr` to bring in data from other tables. `left_join()` implies that you keeping your main table to the left and adding additional columns to the right. In the `by = ` statement you specify which columns are the same so that the join knows what to add where. In the sample code below, we add the description of the different conservation states into our main `msleep` table. The main data contained an extra `domesticated` label which I wanted to keep (done in last line of table with an `ifelse()`)
```{r}
msleep %>% select(name, conservation) %>% 
  mutate(conservation = toupper(conservation)) %>% 
  left_join(conservation_tab, by = c("conservation"="abbreviation")) %>%
  mutate(description = ifelse(is.na(description), conservation, description))
```
## **Spreading and gathering data**
The `gather()` function will gather up many columns into one.In this case we have 3 columns that describe a time measure. For some analysis and graphs, it might be necessary to get them all into one.
The gather function needs you to give a name ("key") for the new descriptive column and another name ("value") for the value column. The columns that you don't want ti gather need to be deselected at the end. In the sample code I'm deselecting the column `name`.
```{r}
msleep %>%
  select(name, contains("sleep")) %>% 
  gather(key = "sleep_measure", value = "time", -name)
```
A useful argument in gathering is the `factor_key` which is `FALSE` by default, but which if set to `TRUE` creates an ordered factor.
```{r}
(msleep_g <- msleep %>%
   select(name, contains("sleep")) %>% 
   gather(key = "sleep_measure", value = "time", -name, factor_key = T)) 
```
The opposite of gathering is spreading. Spread takes one column and makes multiple columns out of it.
```{r}
msleep_g %>%
  spread(key = "sleep_measure", value = "time")
```
## **Turning data into NA **
The function `na_if()` turns particular values into NA. In most cases the command would probably be `na_if("")` (i.e turn an empty string into NA), but it can principally do anything. Here we turn any value that reads "omni" to NA.
```{r}
msleep %>% 
  select(name:order) %>%
  na_if("omni")
```

# **Data Wrangling Part 3: Basic and more advanced ways to filter rows **

## **Basic row filters**

In many cases you don't want to include all rows in your analysis but only a selection of rows. The function to use is dplyr's `filter()`. The general syntax is `filter(dataset, condition)`.
### **Filtering rows based on a numeric variable **
You can filter numeric variables based on their values using operands such as `>`, `>=`, `<=`, `==`, and `!=`.
```{r}
msleep %>% select(name, sleep_total) %>%
  filter(sleep_total > 18)
```
If you want to select a range of values you can use two logical requirements. For instance to select all animals with a total sleep time between 15 and 18 hours, we could use: 
`filter(sleep_total >= 15, sleep_total <=18)`, but there is a shorter and direct way using the `between()` function.
```{r}
msleep %>%
  select(name, sleep_total) %>%
  filter(between(sleep_total, 15,18))
```
Another function that comes handy is `near()`, which will select all code that is nearly a given value. You have to specify a tolerance `tol` to indicate how far the values can be. You can add a specific number: `filter(near(sleep_total,17, tol = 0.5))` will return any rows where `sleep_total` is between 16.5 and 17.5, or you can add a formula. Code below returns rows that are within one standard deviation of 17.
```{r}
msleep %>%
  select(name, sleep_total) %>%
  filter(near(sleep_total, 17, tol = sd(sleep_total)))
```
### **Filtering based on exact character variable matches **
If you want to select a specific group of animals for instance you can use the `==` comparison operator:
```{r}
msleep %>%
  select(order, name, sleep_total) %>%
  filter(order == "Didelphimorphia")
```
Similarly, you can use other operators:
* `filter(order != "Rodentia") will select everything except the Rodentia rows.
*   `filter(name > "v") will select rows with a name in the alaphabet after the letter v.
If you want to select more than one subgroup you can use the `%in%` operator.
```{r}
msleep %>%
  select(order, name, sleep_total) %>%
  filter(order %in% c("Didelphimorphia", "Diprotodontia"))
```
You can use the  `%in%` operator to deselect certain groups as well by adding an exclamation mark at the beginning of your `filter`. Making a `!%in%` may seem logic but won't work.
```{r}
remove <- c("Rodentia", "Carnivora", "Primates")
msleep %>%
  select(order, name, sleep_total) %>%
  filter(!order %in% remove)
```
### **Filtering rows based on regex **
Options above work if you can use the full variable names. In some cases though it will be needed to filter based on partial matches.In this case we need a function that will evaluate regular expressions on strings and return boolean values. Whenever the statement is `TRUE` the row will be filtered. Tere are two options for this: base R's `grepl()` function, or `str_detect()` from `stringr` package.
whwnever you are looking for partial matches, it is important to remember that R is case sensitive. So it is good practice to write something like `filter(grepl(pattern = "mouse", tolower(name)))`.
```{r}
msleep %>% 
  select(name, sleep_total) %>% 
  filter(str_detect(tolower(name), pattern = "mouse"))
```
### **Filtering based on multiple conditions **
The filter option also allows the AND and OR style filters:
* `filter(condition1, condition2)` will return rows where both conditions are met.
* `filter(condition1, !condition2) will return all rows where condition one is true but condition2 is not.
* `filter(condition1|condition2)` will return all rows where condition 1 and/or condition 2 is met.
* `filter(xor(condition1, condition2))` will return all rows where only one of the conditions is met, and not when both conditions are met.
Multiple AND, OR, NOT conditions can be combined.
```{r}
msleep %>%
  select(name, order, sleep_total:bodywt) %>%
  filter(bodywt > 100, (sleep_total > 15 | order != "Carnivora"))
```
Example with `xor()`
```{r}
msleep %>%
  select(name, bodywt:brainwt) %>%
  filter(xor(bodywt > 100, brainwt >1))
```
Example with `!`:
The sample code below will select all rows where `brainwt` is larger than 1, but `bodywt` does not exceed 100.
```{r}
msleep %>%
  select(name, sleep_total, brainwt, bodywt) %>%
  filter(brainwt > 1, !bodywt > 100)
```
### **Filtering out empty rows **
To empty out empty rows, you negate the `is.na()` function inside a filter:
The sample code will remove any rows where `conservation` is `NA`.
```{r}
msleep %>%
  select(name, conservation:sleep_cycle) %>%
  filter(!is.na(conservation))
```
## **Filtering across multiple columns **
The `dplyr` package has a few powerful variants to filter across multiple columns in one go:
* `filter_all()` will filter all columns based on your further instructions
* `filter_if()` requires a function that returns a boolean to indicate which columns to filter on. If that is true, the filter instructions will be followed for those columns.

* `filter_at()` requires you to specify columns inside a `vars()` argument for which the filtering will be done.

### **Filter all **
In the code below I'm searching for the string "Ca" across all columns. I want to keep rows where the string "Ca" is present in ANY of the variables, so I will wrap the condition in `any_vars()` . 
```{r}
msleep %>%
  select(name:order, sleep_total, -vore) %>%
  filter_all(any_vars(str_detect(., pattern = "Ca")))
```
The same can be done for numerical values: This code will retain any rows that has any value below 0.1:
```{r}
msleep %>%
  select(name, sleep_total:bodywt) %>%
  filter_all(any_vars(. < 0.1))
```
The `any_vars()` statement is equivalent to OR, so of course there is an equivalent for AND statements as well: `all_vars()` .  The below code will retain any rows where all values are above 1.
```{r}
msleep %>%
  select(name, sleep_total: bodywt, -awake) %>%
  filter_all(all_vars(. >1))
```
### **Filter if **
So imagine you want to find out all data rows where we have NA in the first few columns. `filter_all(any_vars(is.na(.)))` will be quite useless as it would return 27 rows, many of which are missing data in the measurement section.

n this case: `filter_if()` comes in handy. The describing columns are all character columns while the measurement data is numeric. So using `filter_if()` i can specify that I want to just filter on character variables. In this case I only get 7 rows.
```{r}
msleep %>%
  select(name: order, sleep_total: sleep_rem) %>%
  filter_if(is.character, any_vars(is.na(.)))
```
Similarly, you can add `is.numeric`, `is.integer`, `is.double`, `is.logical`, `is.factor`. If you have date columns you can load the lubridate package and use `is.POSIXt`, `is.Date` .

### **Filter at **
One of the most powerful functions is `filter_at()` : It does not filter all columns, nor does it need you to specify the type of column, you can just select the columns to which the change should happen via the `vars()` argument. This argument allows anything that can be done within a select statement: so you can refer to them by name, but also by by logical numerical functions, regex, etc.
The second argument is the condition for selection. Similar to the examples above, you can use `all_vars()` if all columns need to return TRUE (AND equivalent), or `any_vars()` in case just one variable needs to return TRUE (OR equivalent).

Example: refer to columns by name:
```{r}
msleep %>%
  select(name, sleep_total: sleep_rem, brainwt:bodywt) %>%
  filter_at(vars(sleep_total, sleep_rem), all_vars(. >5))
```

Example: Using another selection option:
```{r}
msleep %>%
  select(name, sleep_total:sleep_rem, brainwt:bodywt) %>%
  filter_at(vars(contains("sleep")), all_vars(. >5))
```
# **Data wrangling Part 4: Summarizing and slicing your data **
In this section we summarize our data: i) counting cases and observations, ii) creating summaries using `summarise()` and it's `summarise_all()`, `_if()` and `_at()` variants, and iii) pulling the maximum and minimum row values.

## **Counting cases and adding counts **
### **Counting the number of observations **
The easiest way to know the number of observations you have for a specific variable, is to use `count()`. By adding the `sort = TRUE` argument, it immediately returns a sortef table with descending number of observations:
```{r}
msleep %>%
  count(order, sort = T)
```
You can add multiple variables to a `count()` statement; the code below counts by order and vore.
```{r}
msleep %>%
  count(order, vore, sort = T)
```
### **Adding the number of observations in a column **
If you're only interested in counting the total num er of cases for a dataframe, you could use `tally()`, which behaves similarly to `nrow()`.
```{r}
msleep %>%
  tally()
msleep %>%
  nrow()
```
More interesting is the `add_tally()` function which automatically adds a colums with the overal number of observations. This would be the same as `mutate(n = n())`.
```{r}
msleep %>%
  select(1:3) %>%
  add_tally()
msleep %>%
  select(1:3) %>%
  mutate(n = n())
```
Even more interesting is the `add_count() ` which takes a variable as an argument, and addds a column with the number of observations. this saves the combination of grouping, mutating and ungrouping again.
```{r}
msleep %>%
  select(name:vore) %>%
  add_count(vore)
```
## **Summarising data **
Both UK and US variants `summarize()` and `summarise()` are applicable within `dplyr` .
If you just want to know the number of observations `count()` does the job, but to produce summaries of the average, sum, standard deviation, minimum, maximum of the data, we need `summarise()`. To use the function you just add your new column name, and after the equal sign the mathematics of what needs happen: `column_name = function(variable). You can add multiple summary functions behind each other.
```{r}
msleep %>%
  summarise(n = n(), average = mean(sleep_total), maximum = max(sleep_total))
```
In most cases, we don't just want to summarise the whole data table, but we want to get summaries by a group. To do this, you first need to specify by which variable(s) you want to divide the dat using `group_by()` . You can add one or more variables as arguments in `group_by()`. 
```{r}
msleep %>%
  group_by(vore) %>%
  summarise(n = n(), average = mean(sleep_total), maximum = max(sleep_total))
```
The `summarise()` call works with nearly any aggregate function, and allows for additional arithmetics:
* `n()` gives the number of observations
* `n_distinct(var)` gives the number of unique values of `var`
* `sum(var)`, `max(var)`, `min(var)` , ...
* `mean(var)`, `median(var)`, `sd(var)`, `IQR(var)` , ...
The code below will average sleep_total and divide by 24, to get the amount of sleep as a fraction of a day.
```{r}
msleep %>%
  group_by(vore) %>%
  summarise(avg_sleep_day = mean(sleep_total)/24)
```
 Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
## **Variants to summarise **
* `summarise_all()` will summarise all columns based on your further instructions
* `summarise_if()` requires a function that returns a boolean, if true the summary instructions will be followed
* `summarise_at()` requires you to specify columns inside a `vars()` argument for which the summary is done.
### **Summarise all **
The function `summarise_all()` requires a function as argument, which it will apply to columns. The sample code calculates the mean for each column. You need to add the `na.rm = T` to remove missing values.
```{r}
msleep %>%
  group_by(vore) %>%
  summarise_all(mean, na.rm = T)
```
The instructions to summarizing have to be a function. When there is no function available in base R or package to do what you want, you can either make a function upfront, or make a function on the fly
The sample code will add 5 to the mean of each column. the function on the fly can be made by either using `funs(mean(., na.rm = T) = 5)` or via a tilde: `~mean(., na.rm = T) + 5` .
```{r}
msleep %>%
  group_by(vore) %>%
  summarise_all(~mean(., na.rm = T) + 5)
```
### **Summarise if **
The function `summarise_if()` requires two arguments:
* Information about columns you want to consider
* Information about how to summarise the data
The sample code below returns the average of all numeric columns
```{r}
msleep %>%
  group_by(vore) %>%
  summarise_if(is.numeric, mean, na.rm = T)
```
One of the downsides of the aggregate summarise functions is that you do not require a new column title. It mya therefore not be clear what the values refer to (average? median? minimum?). You can use `rename_*()` functions, it only takes one extra line of code to accomplish:
```{r}
msleep %>%
  group_by(vore) %>%
  summarise_if(is.numeric, mean, na.rm = T) %>% 
  rename_if(is.numeric, ~paste0("avg_", .))
```
## **Summarise at **
The function `summarise_at()` also requires two arguments:
* First it needs information about the columns you want to consider. In this case you need to wrap them inside a `vars()` statement. Inside `vars()` you can use anything that can be used inside a `select()` statement.
* Secondly, it needs information about how to summarise that data, which as above needs to be a function. If not a function, youi can create a function on the fly using `funs()` or a tilde
The sample code below will return the average of all columns which contain the word "sleep", and also rename them to "avg_var" for clarity.
```{r}
msleep %>%
  group_by(vore) %>%
  summarise_at(vars(contains("sleep")), mean, na.rm = T) %>%
  rename_at(vars(contains("sleep")), ~paste0("avg_",.))
```
## **Arranging rows **
It is useful if your summary tables are arranged, which is `arrange()` 's job. 
Sorting numeric variables:
`arrange(sleep_total)` will arrange it in ascending order. You specify the `desc()` option to arrange in descending order.
```{r}
msleep %>%
  group_by(vore) %>%
  summarise(avg_sleep = mean(sleep_total, na.rm = T)) %>%
  arrange(desc(avg_sleep))
```
If you already grouped your data, you can refer to that group within the `arrange()` statement as well by adding a `.by_group = TRUE` statement. This will sort by descending total sleep time, but within each each group.
```{r}
msleep %>%
  select(order, name, sleep_total) %>%
  group_by(order) %>%
  arrange(desc(sleep_total), .by_group = T)
```
## **Showing only part of your data **
In some cases you don't want to just show all rows available. Here are some nice shortcuts which can save time
### **The 5 lowest and highest values **
If you want to select the highest 5 cases, you could combine an `arrange` call with `head(n=5)`. Or you can use `top_n(5)` which will retain (unsorted) the 5 highest values.
```{r}
msleep %>%
  group_by(order) %>%
  summarise(average = mean(sleep_total)) %>%
  top_n(5)
```
The five lowest values can be found using `top_n(-5)` :
```{r}
msleep %>%
  group_by(order) %>%
  summarise(average = mean(sleep_total)) %>%
  top_n(-5)
```
if you have more than one column, you can add the variable you want it to use. 
```{r}
msleep %>%
  group_by(order) %>%
  summarise(average_sleep = mean(sleep_total), max_sleep =max(sleep_total)) %>% 
  top_n(5, average_sleep)
```
### **A random selection of rows **
using `sample_n()` you can sample a random selection of rows.
Alternative is `sample_frac()` allowing you to randomly select a fraction of rows (here 10 %).
```{r}
msleep %>%
  sample_frac(.1)
```
### **A user-defined slice of rows **
The `head()` call will standard show the first 6 rows, which can be modified by adding an n-argument: `head(n=10)` . Similarly `tail()` will show the final 6 rows, which can also be modified by adding an n-argument. If  you want to slice somewherein the middle, you can use `slice()` . The sample code will show rows 50 to 55.
```{r}
msleep %>%
  slice(50:55)
```