---
title: 'GLM: Generalized Linear Models'
author: "Project Assignment"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
output:
  html_document:
    df_print: paged
  bookdown::pdf_document2: default
  pdf_document: default
  bookdown::word_document2: default
  fig_caption: yes
  word_document: default
tables: yes
fontsize: 14pt
mainfont: Arial
fig_caption: yes
number_sections: yes
toc: yes
header-includes: \usepackage{amsmath}
editor_options:
  chunk_output_type: console
---



```{r setup, child="_setup.Rmd"}
# knitr::opts_chunk$set(echo = TRUE)

```


```{r}
ipk = function(pkg) {
  new.pkg = list.pkg[!(list.pkg %in% .packages(all.available = TRUE))]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE, repos = 'https://cran.us.r-project.org')
  ddpcr::quiet(sapply(pkg, require, character.only = TRUE))
}
list.pkg = c('magrittr',
             'tidyverse',
             'rvest',
             'knitr',
             'httr',
             'jsonlite')

ipk(list.pkg)
```
# Web scraping in R


## HTML Data structure

Web pages are styles with CSS files: cascade style sheets that determine layout of the webpage.
CSS selectors can be used to look for HTML elements of interest. One such is the [SelectorGadget google chrome extension](https://selectorgadget.com/). You need to install it to your browser before proceeding.

To use it open the page 

1. Click on the element you want to select. SelectorGadget will make a first guess at what css selector you want. It's    likely to be bad since it only has one example to learn from, but it's a start. Elements that match the selector      will be highlighted in yellow.

2. Click on elements that shouldn't be selected. They will turn red. Click on elements that should be selected. They     will turn green.

3. Iterate until only the elements you want are selected. SelectorGadget isn't perfect and sometimes won't be able to    find a useful css selector. Sometimes starting from a different element helps. More at [tidyverse/rvest](https://rvest.tidyverse.org/articles/selectorgadget.html)

For example, if we want the actors listed on the IMDB movie page, e.g. [The Shawshank Redemption](https://www.imdb.com/title/tt0111161/?ref_=adv_li_tt)


HTML tags can be passed to functions to retrieve the web page elements of interest.

## `rvest`

For scrapping (harvesting) data fro the web in a structured format that can be used in further analysis.

### rvest functions

- `read_html()`: collects data from the webpage
- `html_nodes()`: extract the relevant pieces
- `html_text()`: extract tags of the relevant piece
- `html_attributes()`: extract attributes of the relevant piece

```{r}
# specify url 
url = 'https://www.imdb.com/title/tt0111161/'
# reading the html code from the 
webpage = read_html(url)
webpage
```

Once we have determined the CSS selector, we use it to extract the information we want

```{r}
cast = html_nodes(webpage, ".primary_photo+ td a")
length(cast)
cast[1:2]
```

Finally, we extract the text from the selected HTML nodes.

```{r}
html_text(cast, trim = T)
```

#### Extracting a table

```{r}
all_tables = html_table(webpage, "table", header = FALSE)
casttable = html_table(webpage, ".cast_list", header = F)[[1]]
head(casttable)
```


#### Attributes of an element

If say we are also interested in extracting the links to the actor's pages, we can acces html attributes of the selected nodes using `html_attrs( )`.

```{r}
cast_attrs = html_attrs(cast)
cast_attrs[1:2]
```

As we can see there's only one attribute called `href` which contains relative url to the actor's page. We can extract it using `html_attr()`, indicating the name of the attribute of interest. Relative urls can be turned into absolute urls using `url_absolute()`.

```{r}
cast_rel_urls = html_attr(cast, "href")
length(cast_rel_urls)
cast_rel_urls[1:2]
cast_abs_urls = html_attr(cast, "href") %>%
  url_absolute(url)
cast_abs_urls[1:2]
```


## Making API Requests in R

- Application Program Interface: Places where a computer interacts with another, or with itself:
  - the client requests data, the server provide data.
- Applicable R packages: `httr` and `jsonlite`

### `httr`

- Create request with `GET()` function. Input is a url which specifies the address of the server.

- Example: current number of people in space

```{r}
res = GET('http://api.open-notify.org/astros.json')
res
```

#### JSON Format

- Different formats to share data on the internet
- Currently, JavaScript Object notation is being widely adopted



[
{
"name": "Miguel",
"student_id": 1,
"exam_1": 85,
"exam_2": 86
},
{
"name": "Sofia",
"student_id": 2,
"exam_1": 94,
"exam_2": 93

### jsonlite

- Several organizations provide a JSON API or a web service
- The `jsonlite` package provides parser and generator functions:
    - `fromJSON()`   
    - `toJSON()`

- `fromJSON()`:
  - Input is a JSON string, URL, JSON file
  - Returns a list of data.frames

- `toJSON()`:
  - Input is any object
  - Returns a JSON string


#### `fromJSON()`

- convert the raw Unicode of the GET request into a JSON string

```{r}
rawToChar(res$content)
```

- convert the JSON format to a list of data.frames

```{r}
data = fromJSON(rawToChar(res$content))
data
```

Read directly from a url

- for example: read from a github page


```{r}
data <- fromJSON("https://api.github.com/users/hadley/orgs")
class(data)
names(data)
```


#### `toJSON()`

- Write a dataframe to JSON format

```{r}
jsondata = toJSON(data)
head(jsondata)
```

And back from JSON format to dataframe

```{r}
backagain = fromJSON(jsondata)
identical(data, backagain)
```


#### API with Query Parameters

- When is the ISS was going to pass over a given location on earth?

**The ISS and the Brooklynn Bridge**

- Query parameters:
  - longitude
  - latitude
- Combine with the original URL

```{r}
res = GET("http://api.open-notify.org/iss-pass.json",
query = list(lat = 40.7, lon = -74))
res
```


```{r}
data = fromJSON(rawToChar(res$content))
data$response
```


- Check the documentation of the API you are using
  - required and optional parameters
  - authentication

- Not every API works with query parameters
  - Note the “?{var}=” construction in the URL of the ISS
  - Study the URL before building your function



